---
title: "Let's Practice Geovisualization and Spatial Analysis "
author: "Kim Antunez & Timothée Giraud"
date: "Massive spatial data: challenges in acquisition, treatment and use for territories"
output:
  unilur::tutorial_html_solution:
    toc: true
    toc_float: false
    toc_depth: 1
    suffix: ""
    theme: journal
    highlight: kate
    number_sections: no
    number_subsections: no
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
library(stringr)
library(knitr)
## Global options
options(max.print="90")
opts_chunk$set(echo=TRUE,
	             cache=FALSE, #TRUE
               prompt=FALSE,
               tidy=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=90)
options(width = 90)

# no margins
knit_hooks$set(nm = function(before, options, envir){
  if (before){
    par(mar=c(0,0,0,0))
  }
})

# title margins
knit_hooks$set(sm = function(before, options, envir){
  if (before){
    par(mar=c(0,0,1.2,0))
  }
})

# boxes custom
#remotes::install_github("koncina/unilur")
knitr::opts_template$set(alert = list(box.title = "Watch out!",
                                      box.body = list(fill = "#ffe0d9", colour = "black"),
                                      box.header = list(fill = "#FFAD99", colour = "black"),
                                      box.collapse = NULL,
                                      box.icon = "fa-exclamation-triangle"))
knitr::opts_template$set(solution = list(box.title = "Solution",
                                         box.body = list(fill = "#e6f6e7", colour = "black"),
                                         box.header = list(fill = "#ace1af", colour = "black"),
                                         box.icon = "fa-check-square",
                                         box.collapse = TRUE))
knitr::opts_template$set(information = list(box.title = "Information",
                                            box.body = list(fill = "#bbe8f4", colour = "black"),
                                            box.header = list(fill = "#64c9e6", colour = "black"),
                                            box.icon = "fa-info-circle",
                                            box.collapse = NULL))
knitr::opts_template$set(clues = list(box.title = "Clues",
                                      box.body = list(fill = "#fff9dc", colour = "black"),
                                      box.header = list(fill = "#ffec8b", colour = "black"),
                                      box.icon = "fa-search",
                                      box.collapse = TRUE))
```

The main data used in this tutorial and in the lecture are about the **geolocalisation of french restaurants** in Paris and in a department called Haute-Garonne. We use two different sources: 

1. an official register called [**SIRENE**](https://www.data.gouv.fr/fr/datasets/base-sirene-des-entreprises-et-de-leurs-etablissements-siren-siret/) (Computer system for the business and establishment register) managed by the French National Institute of Statistics and Economic Studies ([Insee](https://www.insee.fr/en/accueil)) and [geolocated](http://data.cquest.org/geo_sirene/last/) by [Etalab](https://www.etalab.gouv.fr/) (French task force for Open Data). This register records the civil status of all companies and their establishments (including restaurants).

SIRENE has the advantages of being rigorous and exhaustive on the French territory.

2. the famous global open access mapping project [**OpenStreetMap**](https://www.openstreetmap.org/) (OSM). We can access to OSM database using the [osmdata](https://github.com/ropensci/osmdata) R package.

OSM has many benefits, ensuring transparent data provenance and ownership, enabling real-time evolution of the database and, by allowing anyone to contribute, encouraging democratic decision making and citizen science. 
 


# Exercise 1 : Manipulate sf objects and associated data.frames

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
1. Load in R the iris[^1] map layer ‘iris_31.rds’ of the french department called Haute-Garonne (numbered 31). Would the `sf::st_read` function also work? Why?
```


[^1]: In French, IRIS is an acronym of ‘aggregated units for statistical information’. Their target sizes are 2000 residents per basic unit.

```{block, opts.label = "clues"}
Use the `readRDS` function.
```

```{r, solution = TRUE}
library(sf)
iris_31 <- readRDS("data/iris_31.rds")
# iris_31 <- st_read("data/iris_31.rds")
```

```{block, solution = TRUE}
`st_read` would not work because ‘iris_31.rds’ is not a shapefile but a file already R formatted. Simply load it with the `readRDS` function.
```


```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Display the basemap of department 31 with `plot(iris_31)`.
What do you notice ?
```

```{r, nm=TRUE, solution = TRUE}
plot(iris_31)
```

```{block, solution = TRUE}
We notice that R performs 3 graphs: one graph per variable in the sf object.
```

```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
What is the functionality of the `sf::st_geometry` function? What solution do you propose then?
```

```{block, solution = TRUE}
`sf::st_geometry` makes it possible to isolate the information contained in the ‘geometry’ column of the sf object. Using it, we put aside other variables (here *CODE_IRIS*, *P14_POP* and *INSEE_COM*).
```

```{r, nm=TRUE, solution = TRUE}
# ?sf::st_geometry
plot(st_geometry(iris_31))
```


```{block, box.title = "4", box.body = list(fill = "white"), box.icon = "fa-star"}
In which projection is the map layer? Map it with another projection.
```

```{block, opts.label = "information"}
Test the Azimuthal Equidistant projection with "`crs="+proj=aeqd +lat_0=90 +lon_0=0`" to see a clear difference and create a layer called ‘iris_31_aeqd’.
```

```{block, opts.label = "clues"}
Use the `sf::st_crs` function to guess the projection and `sf::st_transform` to change the projection.
```

```{r, nm=TRUE, eval=FALSE, solution = TRUE}
#?st_crs
st_crs(iris_31)
par(mar = c(0,0,0,0), mfrow = c(1,2))
plot(st_geometry(iris_31))
iris_31_aeqd <- st_transform(iris_31, crs="+proj=aeqd +lat_0=90 +lon_0=0")
plot(st_geometry(iris_31_aeqd))
```

```{r, nm=TRUE, echo=FALSE}
par(mar = c(0,0,0,0), mfrow = c(1,2))
plot(st_geometry(iris_31))
iris_31_aeqd <- st_transform(iris_31, crs="+proj=aeqd +lat_0=90 +lon_0=0")
plot(st_geometry(iris_31_aeqd))
```

```{block, box.title = "5", box.body = list(fill = "white"), box.icon = "fa-star"}
Calculate the distance matrix between the 5 first iris of department 31. Do you get the same distance matrix if you are working on a layer projected in another projection?
```

```{block, opts.label = "information"}
Use map layers called ‘iris_31’ and ‘iris_31_aeqd’.
```

```{r, nm=TRUE, solution=TRUE}
mat <- st_distance(x = iris_31[1:5,], y = iris_31[1:5,])
mat

mat_aeqd <- st_distance(x = iris_31_aeqd[1:5,], y = iris_31_aeqd[1:5,])
mat_aeqd

identical(mat,mat_aeqd)
```


```{block, solution = TRUE}
No, the two matrices are different.
```

```{block, box.title = "6", box.body = list(fill = "white"), box.icon = "fa-star"}
Using the layer called ‘iris_31’, create a new aggregated map layer called ‘com_31’ which corresponds to the municipalities of department 31. Also keep in this new layer the information on the population in each municipality.
```

```{block, opts.label = "information"}
The map layer called ‘iris_31’ contains the 5 digit codes of municipalities in its variable *INSEE_COM* and the 2014 population in its column *P14_POP*.
```

```{block, opts.label = "clues"}
Use the classic functions of `dplyr` package: `select`, `group_by` et `summarize`.
These functions also work with `sf` objects.
```


```{r, nm=TRUE, eval=FALSE, solution = TRUE}
library(dplyr)
com_31 <- iris_31 %>%
  select(INSEE_COM,P14_POP) %>%
  group_by(INSEE_COM) %>% 
  summarize(P14_POP= sum(P14_POP)) %>% 
  st_cast("MULTIPOLYGON")

plot(st_geometry(com_31))
```

```{r, nm=TRUE, echo=FALSE}
library(dplyr)
com_31 <- iris_31 %>%
  select(INSEE_COM,P14_POP) %>%
  group_by(INSEE_COM) %>% 
  summarize(P14_POP= sum(P14_POP)) %>% 
  st_cast("MULTIPOLYGON")

plot(st_geometry(com_31))
```

```{block, box.title = "7", box.body = list(fill = "white"), box.icon = "fa-star"}
Using the data contained in ‘sir_31’, add to this layer the number of restaurants per municipality. 
```

```{block, opts.label = "information"}
The code of each municipality is not in the ‘sir_31’ database. To create it, you have to create a variable called *INSEE_COM* (5 digits) which concatenates the *DEPET* (2 digits) and *COMET* (3 digits) variables.
```

```{r, nm=TRUE, solution = TRUE}
sir_31 <- readRDS("data/sir_31.rds")

com_31 <- left_join(com_31,
                    sir_31 %>%
                      mutate(INSEE_COM=paste0(DEPET,COMET)) %>% 
                      group_by(INSEE_COM) %>% 
                      summarize(nb_of_rest= n()) %>% 
                      st_set_geometry(NULL),
                    by=c("INSEE_COM"="INSEE_COM")) %>% 
  mutate(nb_of_rest=ifelse(is.na(nb_of_rest),0,nb_of_rest))
```

```{block, box.title = "8", box.body = list(fill = "white"), box.icon = "fa-star"}
Aggregate all the information present in ‘com_31’ at the level of french intercommunalites (called EPCI) and call this new layer ‘epci_31’. 
```


```{block, opts.label = "information"}
You have to use the database ‘table_MAUP.rds’ to have a match between the municipality code (*CODGEO*) and intercommunality code (*EPCI*).
```


```{r, nm=TRUE, eval=FALSE, solution = TRUE}
table_MAUP <- readRDS("data/table_MAUP.rds") %>% 
  select(CODGEO,EPCI)

epci_31  <- com_31 %>%
  left_join(table_MAUP,by=c("INSEE_COM"="CODGEO")) %>% 
  group_by(EPCI) %>% 
  summarize(P14_POP=sum(P14_POP),nb_of_rest= sum(nb_of_rest)) %>% 
  st_cast("MULTIPOLYGON")

plot(st_geometry(epci_31))
```


```{r, nm=TRUE, echo=FALSE}
table_MAUP <- readRDS("data/table_MAUP.rds") %>% 
  select(CODGEO,EPCI)

epci_31  <- com_31 %>%
  left_join(table_MAUP,by=c("INSEE_COM"="CODGEO")) %>% 
  group_by(EPCI) %>% 
  summarize(P14_POP=sum(P14_POP),nb_of_rest= sum(nb_of_rest)) %>% 
  st_cast("MULTIPOLYGON")

plot(st_geometry(epci_31))
```



```{block, box.title = "9", box.body = list(fill = "white"), box.icon = "fa-star"}
Using the `cartography` package, simply plot a map of french intercommunality with a  proportional circle layer related to the number of restaurants.
```

```{block, opts.label = "clues"}
The `propSymbolsLayer` function allows you to draw proportional circles.
```

```{r, nm=TRUE, eval=FALSE, solution = TRUE}
library(cartography)
plot(st_geometry(epci_31), col = "ivory1", border = "ivory3",lwd =0.5,bg = "#FBEDDA")
propSymbolsLayer(epci_31, var = "nb_of_rest", inches = 0.2)
```

```{r, nm=TRUE, echo=FALSE}
library(cartography)
plot(st_geometry(epci_31), col = "ivory1", border = "ivory3",lwd =0.5,bg = "#FBEDDA")
propSymbolsLayer(epci_31, var = "nb_of_rest", inches = 0.2)
```

```{r, nm=TRUE, echo=FALSE, eval=FALSE}
#Saving objects for following exercises
#saveRDS(object = epci_31, file = "data/epci_31.rds")
#saveRDS(object = com_31, file = "data/com_31.rds")
```

# Exercise 2 : Maps with cartography and ggplot2 packages

We would like here to design EPCI maps that combine the number of restaurants and the number of restaurants per 10,000 inhabitants. 



```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Data preparation:

* Load ‘fra.shp’ in R: the map layer of the borders of the French metropolitan territory. 
* Load the layer called ‘epci_31.rds’ (see previous exercise for more details) and 
create a variable called *nb_rest_10000inhab* which corresponds to the number of restaurants per 10,000 inhabitants in each territory.
* Create a vector of quantiles breaks of the *nb_rest_10000inhab* variable.
* Create the vector colors which corresponds to a the number of classes defined earlier.
* Finally, add  a variable called *typo* to 'epci_31' which indicates the class of the territory according to the discretization contained in *bks* for the  *nb_rest_10000inhab* variable (this is needed for ggplot2 maps).
```

```{block, opts.label = "information"}
For the creation of ‘bks’ et ‘cols’, use the `getBreaks` et `carto.pal` functions of the  `cartography` package.
For the creation of the *typo* variable, you can use the `cut` function and apply the parameters `digit.lab = 2` and `include.lowest = TRUE`.
```

```{r, solution=TRUE}
library(sf)
library(cartography)
library(dplyr)
# Import data
fra <- st_read("data/fra.shp", quiet = TRUE)
epci_31 <- readRDS("data/epci_31.rds")
# Create the variable
epci_31$nb_rest_10000inhab <- 10000 * epci_31$nb_of_rest / epci_31$P14_POP
# Define breaks
bks <- getBreaks(v = epci_31$nb_rest_10000inhab, method = "quantile", nclass = 4)
# Define color palette
cols <- carto.pal("orange.pal", length(bks)-1)
# Create a "typo"" variable
epci_31 <- epci_31 %>%
  mutate(typo = cut(nb_rest_10000inhab,breaks = bks, dig.lab = 2, 
                    include.lowest = TRUE))
```


```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
With the help of `cartography` package, make the following map which contains in a choropleth layer the variable *nb_rest_10000inhab* and in a proportional circle layer the variable *nb_of_rest*. Do the same thing with the `ggplot2` package.
```


With cartography:

```{r, eval=FALSE, solution=TRUE}
# Define plot margins
par(mar = c(0.2, 0.2, 1.4, 0.2), bg = "azure")
# Find EPCI bounding box
bb <- st_bbox(epci_31)
# Plot France using EPCI boundingbox
plot(st_geometry(fra), col="ivory", border = "ivory3", 
     xlim = bb[c(1, 3)], ylim = bb[c(2, 4)])
# Plot the choropleth layer
choroLayer(epci_31, var = "nb_rest_10000inhab", 
           breaks = bks, col = cols, border = "grey80", lwd = 0.5, 
           legend.pos = "topleft",add = TRUE,
           legend.title.txt = "Number of restaurants\nfor 10,000 inhabitants")
# Plot proportionnal symbols
propSymbolsLayer(epci_31, var="nb_of_rest", col="#440170",border=NA, 
                 legend.pos="left", inches=0.4, add = TRUE,
                 legend.title.txt = "Number of restaurants")
# Add a layout layer
layoutLayer(title = "Restaurants", sources = "Insee, 2018", 
            author = "Kim & Tim, 2018", 
            theme = "green.pal", col = "darkred", 
            coltitle = "white", postitle = "center",
            frame = TRUE, scale = 10)
# Add a north (south) arrow
north(pos = "topright", south = TRUE)
```

With ggplot2:

```{r, eval=FALSE, solution=TRUE}
library(ggplot2)

map_ggplot <- ggplot() +
  geom_sf(data = fra, colour = "ivory3",
          fill = "ivory") +
  geom_sf(data = epci_31, aes(fill = typo), colour = "grey80") +
  scale_fill_manual(name = "Number of restaurants\nfor 10,000 inhabitants",
                    values = cols, na.value = "#303030")+
   geom_sf(data = epci_31 %>%  st_centroid(),
           aes(size= nb_of_rest), color = "#440154CC", show.legend = 'point')+
  scale_size(name = "Number of restaurants",
             breaks = c(1, 500, 3200),
             range = c(0,18))+
  coord_sf(crs = 2154, datum = NA,
           xlim = st_bbox(epci_31)[c(1,3)],
           ylim = st_bbox(epci_31)[c(2,4)]
  ) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "azure",color=NA)) +
  labs(
      title = "Restaurants",
      caption = "Insee, 2018\nKim & Tim, 2018"
  )

plot(map_ggplot)
```



#### {.tabset} 

##### cartography
```{r, nm=TRUE, echo=FALSE}
# Define plot margins
par(mar = c(0.2, 0.2, 1.4, 0.2), bg = "azure")
# Find EPCI bounding box
bb <- st_bbox(epci_31)
# Plot France using EPCI boundingbox
plot(st_geometry(fra), col="ivory", border = "ivory3", 
     xlim = bb[c(1, 3)], ylim = bb[c(2, 4)])
# Plot the choropleth layer
choroLayer(epci_31, var = "nb_rest_10000inhab", 
           breaks = bks, col = cols, border = "grey80", lwd = 0.5, 
           legend.pos = "topleft",add = TRUE,
           legend.title.txt = "Number of restaurants\nfor 10,000 inhabitants")
# Plot proportionnal symbols
propSymbolsLayer(epci_31, var="nb_of_rest", col="#440170",border=NA, 
                 legend.pos="left", inches=0.4, add = TRUE,
                 legend.title.txt = "Number of restaurants")
# Add a layout layer
layoutLayer(title = "Restaurants", sources = "Insee, 2018", 
            author = "Kim & Tim, 2018", 
            theme = "green.pal", col = "darkred", 
            coltitle = "white", postitle = "center",
            frame = TRUE, scale = 10)
# Add a north (south) arrow
north(pos = "topright", south = TRUE)
```

##### ggplot2
```{r, nm=TRUE, echo=FALSE}
library(ggplot2)

map_ggplot <- ggplot() +
  geom_sf(data = fra, colour = "ivory3",
          fill = "ivory") +
  geom_sf(data = epci_31, aes(fill = typo), colour = "grey80") +
  scale_fill_manual(name = "Number of restaurants\nfor 10,000 inhabitants",
                    values = cols, na.value = "#303030")+
   geom_sf(data = epci_31 %>%  st_centroid(),
           aes(size= nb_of_rest), color = "#440154CC", show.legend = 'point')+
  scale_size(name = "Number of restaurants",
             breaks = c(1, 500, 3200),
             range = c(0,18))+
  coord_sf(crs = 2154, datum = NA,
           xlim = st_bbox(epci_31)[c(1,3)],
           ylim = st_bbox(epci_31)[c(2,4)]
  ) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "azure",color=NA)) +
  labs(
      title = "Restaurants",
      caption = "Insee, 2018\nKim & Tim, 2018"
  )
plot(map_ggplot)
```


#### {}


```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
What other solution could we use to display these two variable on the same map? Try it using the `cartography` package.
```

```{block, opts.label = "clues"}
The `propSymbolsChoroLayer` function allows you to draw colored proportional circles.
```

```{r, nm=TRUE, eval=FALSE, echo=TRUE, solution=TRUE}
# Define plot margins
par(mar = c(0.2, 0.2, 1.4, 0.2), bg = "azure")
# Find EPCI bounding box
bb <- st_bbox(epci_31)
# Plot France using EPCI boundingbox
plot(st_geometry(fra), col="ivory", border = "ivory3", 
     xlim = bb[c(1, 3)], ylim = bb[c(2, 4)])
# Plot EPCI
plot(st_geometry(epci_31), col="ivory3", border = "ivory2", add=T)
# Plot the choropleth layer
propSymbolsChoroLayer(epci_31, var = "nb_of_rest", var2 = "nb_rest_10000inhab", 
           breaks = bks, col = cols, border = "grey80", lwd = 0.5, 
           legend.var.pos = "topleft", legend.var2.pos = "left",
           add = TRUE, inches = 0.4,
           legend.var.title.txt = "Number of restaurants",
           legend.var2.title.txt = "Number of restaurants\nfor 10,000 inhabitants")
# Add a layout layer
layoutLayer(title = "Restaurants", sources = "Insee, 2018", 
            author = "Kim & Tim, 2018", 
            theme = "green.pal", col = "darkred", 
            coltitle = "white", postitle = "center",
            frame = TRUE, scale = 10)
# Add a north (south) arrow
north(pos = "topright", south = TRUE)

```

```{r, nm=TRUE, echo=FALSE}
# Define plot margins
par(mar = c(0.2, 0.2, 1.4, 0.2), bg = "azure")
# Find EPCI bounding box
bb <- st_bbox(epci_31)
# Plot France using EPCI boundingbox
plot(st_geometry(fra), col="ivory", border = "ivory3", 
     xlim = bb[c(1, 3)], ylim = bb[c(2, 4)])
# Plot EPCI
plot(st_geometry(epci_31), col="ivory3", border = "ivory2", add=T)
# Plot the choropleth layer
propSymbolsChoroLayer(epci_31, var = "nb_of_rest", var2 = "nb_rest_10000inhab", 
           breaks = bks, col = cols, border = "grey80", lwd = 0.5, 
           legend.var.pos = "topleft", legend.var2.pos = "left",
           add = TRUE, inches = 0.4,
           legend.var.title.txt = "Number of restaurants",
           legend.var2.title.txt = "Number of restaurants\nfor 10,000 inhabitants")
# Add a layout layer
layoutLayer(title = "Restaurants", sources = "Insee, 2018", 
            author = "Kim & Tim, 2018", 
            theme = "green.pal", col = "darkred", 
            coltitle = "white", postitle = "center",
            frame = TRUE, scale = 10)
# Add a north (south) arrow
north(pos = "topright", south = TRUE)
```



```{block, box.title = "4", box.body = list(fill = "white"), box.icon = "fa-star"}
Using the `cartography` package, display the number of restaurants and the number of restaurants per 10,000 inhabitants at the municipalities and EPCI scales. The two maps displayed side by side should be as much comparable as possible.
```


```{r, nm=TRUE, eval=TRUE, echo=TRUE, solution=TRUE, fig.width=10}
library(sf)
library(cartography)
library(dplyr)
# Import data
fra <- st_read("data/fra.shp", quiet = TRUE)
epci_31 <- readRDS("data/epci_31.rds")
com_31 <- readRDS("data/com_31.rds")

# Create the variable
epci_31$nb_rest_10000inhab <- 10000 * epci_31$nb_of_rest / epci_31$P14_POP
com_31$nb_rest_10000inhab <- 10000 * com_31$nb_of_rest / com_31$P14_POP

# Define breaks for municipalities (we will use the same breaks for both maps)
bks_com <- getBreaks(v = com_31$nb_rest_10000inhab[com_31$nb_of_rest>0], 
                     method = "quantile", nclass = 6)
# Define color palette
cols <- carto.pal("wine.pal", length(bks_com)-1)

# Define plot margins
par(mar = c(0, 0.1, 1.2, 0.1), bg = "azure", mfrow = c(1,2))
# Find EPCI bounding box
bb <- st_bbox(epci_31)
# Plot France using EPCI boundingbox
plot(st_geometry(fra), col="ivory", border = "ivory3", 
     xlim = bb[c(1, 3)], ylim = bb[c(2, 4)])
# Plot EPCI
plot(st_geometry(epci_31), col="ivory3", border = "ivory2", add=T)
# Plot the choropleth layer
propSymbolsChoroLayer(epci_31, var = "nb_of_rest", var2 = "nb_rest_10000inhab", 
           breaks = bks_com, col = cols, border = "ivory3",lwd = 0.6, 
           legend.var.pos = "bottomright", legend.var2.pos = "n",
           add = TRUE, inches = 0.5,
           legend.var.title.txt = "Number of restaurants")
# Add a layout layer
layoutLayer(title = "Restaurants", sources = "Insee, 2018", 
            author = "Kim & Tim, 2018", 
            theme = "green.pal", col = "darkred", 
            coltitle = "white", postitle = "center",
            frame = FALSE, scale = NULL)


bb <- st_bbox(epci_31)
# Plot France using EPCI boundingbox
plot(st_geometry(fra), col="ivory", border = "ivory3", 
     xlim = bb[c(1, 3)], ylim = bb[c(2, 4)])
# Plot EPCI
plot(st_geometry(com_31), col="ivory3", border = "ivory2",lwd = .5, add=T)
# Plot the choropleth layer
propSymbolsChoroLayer(com_31, var = "nb_of_rest", var2 = "nb_rest_10000inhab", 
           breaks = bks_com, col = cols, border = "ivory3",lwd = 0.6, 
           fixmax = max(epci_31$nb_of_rest),
           legend.var.pos = "n", legend.var2.pos = "bottom",
           add = TRUE, inches = 0.5, legend.var2.values.rnd = 0,
           legend.var2.title.txt = "Number of restaurants\nfor 10,000 inhabitants")
# Add a layout layer
layoutLayer(title = "Restaurants", sources = "", 
            author = "", 
            theme = "green.pal", col = "darkred", 
            coltitle = "white", postitle = "center",
            frame = FALSE, scale = 10)
# Add a north (south) arrow
north(pos = "topright", south = TRUE)


```










# Exercise 3 : Smooth Density analysis of restaurants in Haute-Garonne

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Load the dataset ‘sir_31’ used previously and map the more than 4,000 restaurants of department 31 with the `mapview` package. Try using different parameters to customize your map.
```

```{block, opts.label = "information"}
For example, you can use the `map.types`, `col.regions`, `label`, `color`, `legend`, `layer.name`, `homebutton`, `lwd` ... parameters of the `mapview` function.
```

```{r, nm=TRUE, eval=FALSE, echo=TRUE, solution=TRUE}
library(mapview)
library(sf)
library(cartography)

sir_31 <- readRDS("data/sir_31.rds")
mapview(sir_31, map.types = "OpenStreetMap", 
        col.regions = "#940000", 
        label = paste(sir_31$L2_NORMALISEE, sir_31$NOMEN_LONG, sep = " - "), 
        color = "white", legend = TRUE, layer.name = "Restaurants in SIRENE",
        homebutton = FALSE, lwd = 0.5) 
```

```{r, nm=TRUE, eval=TRUE, echo=FALSE}
library(mapview)
library(sf)
library(cartography)

sir_31 <- readRDS("data/sir_31.rds")
mapview(sir_31, map.types = "OpenStreetMap", 
        col.regions = "#940000", 
        label = paste(sir_31$L2_NORMALISEE, sir_31$NOMEN_LONG, sep = " - "), 
        color = "white", legend = TRUE, layer.name = "Restaurants in SIRENE",
        homebutton = FALSE, lwd = 0.5) 
```


```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Load the layer ‘iris_31’ used previously. Apply the `pt_in_grid` function below on the SIRENE data and on the iris of the department 31 to create a grid called ‘grid’ and count points in it. Use a 2km cellsize. 
```


```{r, opts.label = "information"}
pt_in_grid <- function(feat, adm, cellsize = 50){
  grid <- st_make_grid(x = adm, cellsize = cellsize, what = "polygons")
  . <- st_intersects(grid, adm)
  grid <- grid[sapply(X = ., FUN = length)>0]
  . <- st_intersects(grid, feat)
  grid <- st_sf(n = sapply(X = ., FUN = length), grid)
  return(grid)
}
```

```{r, sm=TRUE, solution=TRUE}
# import
iris_31 <- readRDS("data/iris_31.rds")
# create the grid
grid <- pt_in_grid(sir_31, iris_31, 2000)
```


```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
Simply use the `plot` instruction to plot the grid. Then, using the `choroLayer` function of the `cartography` package, plot the number of restaurants per kilometer (called ‘n’) in each cells of the grid. What observation can you make concerning the readability of the map?
```

```{r, sm=TRUE, echo=TRUE, eval=TRUE, solution=TRUE}
# default plot
plot(grid)
```

```{r, sm=TRUE, echo=TRUE, eval=FALSE, solution=TRUE}
# n = number of restaurants / km2
grid$n <- grid$n*1000*1000/as.numeric(st_area(grid))

# Custom plot with equal length classes
bks <- getBreaks(grid$n[grid$n>0], method = "equal", nclass = 9)
cols <- carto.pal("wine.pal", length(bks)-1)
# or
# cols <- mapview::mapviewGetOption("raster.palette")(length(bks)-1)
opar <- par(mar = c(0,0,1.2,0), bg = "#FBEDDA")
plot(st_geometry(iris_31), col = "ivory3", border = "NA", lwd = 0.5)
choroLayer(grid, var = "n", border = NA,
           breaks = bks, col = cols, legend.values.rnd = 2,
           legend.pos = "topleft", legend.title.txt = "Restaurants / km2",
           add=T)
layoutLayer(title = "Restaurant Distribution in Haute-Garonne", scale = 5, 
            frame = FALSE, tabtitle = TRUE, north = TRUE,
            author = "Kim & Tim", sources = "INSEE - 2018, Sirene - 2018")
```

```{r, sm=TRUE, echo=FALSE, eval=TRUE}
# n = number of restaurants / km2
grid$n <- grid$n*1000*1000/as.numeric(st_area(grid))

# Custom plot with equal length classes
bks <- getBreaks(grid$n[grid$n>0], method = "equal", nclass = 9)
cols <- carto.pal("wine.pal", length(bks)-1)
# or
# cols <- mapview::mapviewGetOption("raster.palette")(length(bks)-1)
opar <- par(mar = c(0,0,1.2,0), bg = "#FBEDDA")
plot(st_geometry(iris_31), col = "ivory3", border = "NA", lwd = 0.5)
choroLayer(grid, var = "n", border = NA,
           breaks = bks, col = cols, legend.values.rnd = 2,
           legend.pos = "topleft", legend.title.txt = "Restaurants / km2",
           add=T)
layoutLayer(title = "Restaurant Distribution in Haute-Garonne", scale = 5, 
            frame = FALSE, tabtitle = TRUE, north = TRUE,
            author = "Kim & Tim", sources = "INSEE - 2018, Sirene - 2018")

```


```{block, sm=TRUE, solution=TRUE}
The map is not very readable because of the choice of the discretization.
```

```{block, box.title = "4", box.body = list(fill = "white"), box.icon = "fa-star"}
Plot the distribution of the number of restaurants per box in the grid. What do you notice and what solution can you suggest? 
```

```{r, eval=TRUE, echo=TRUE, solution=TRUE}
hist(grid$n[grid$n>0], 150)
```

```{block, sm=TRUE, eval=TRUE, echo=TRUE, solution=TRUE}
The distribution is highly dissymetric. We could use instead a geometric progression classification.
```

```{block, box.title = "5", box.body = list(fill = "white"), box.icon = "fa-star"}
Try to make a more readable map.
```

```{block, opts.label = "clues"}
Test the `method = 'geom'` and `nclass = 9` parameters of the `cartography::getBreaks` function
```

```{r, sm=TRUE, eval=FALSE, echo=TRUE, solution=TRUE}
bks <- getBreaks(grid$n[grid$n>0], method = "geom", nclass = 9)
cols <- carto.pal("wine.pal", length(bks)-1)
# or 
# cols <- mapview::mapviewGetOption("raster.palette")(length(bks)-1)
par(mar = c(0,0,1.2,0), bg = "#FBEDDA")
plot(st_geometry(iris_31), col = "ivory3", border = "NA", lwd = 0.5)
choroLayer(grid, var = "n", border = NA,
           breaks = bks, col = cols,
           legend.pos = "topleft", legend.values.rnd = 2,
           legend.title.txt = "Restaurants / km2",
           add=T)
layoutLayer(title = "Restaurant Distribution in Haute-Garonne", scale = 5, 
            frame = FALSE, tabtitle = TRUE, north = TRUE,
            author = "Kim & Tim", sources = "INSEE - 2018, Sirene - 2018")
```


```{r, sm=TRUE, echo=FALSE, eval=TRUE}
bks <- getBreaks(grid$n[grid$n>0], method = "geom", nclass = 9)
cols <- carto.pal("wine.pal", length(bks)-1)
# or 
# cols <- mapview::mapviewGetOption("raster.palette")(length(bks)-1)
par(mar = c(0,0,1.2,0), bg = "#FBEDDA")
plot(st_geometry(iris_31), col = "ivory3", border = "NA", lwd = 0.5)
choroLayer(grid, var = "n", border = NA,
           breaks = bks, col = cols,
           legend.pos = "topleft", legend.values.rnd = 2,
           legend.title.txt = "Restaurants / km2",
           add=T)
layoutLayer(title = "Restaurant Distribution in Haute-Garonne", scale = 5, 
            frame = FALSE, tabtitle = TRUE, north = TRUE,
            author = "Kim & Tim", sources = "INSEE - 2018, Sirene - 2018")
```


```{block, box.title = "6", box.body = list(fill = "white"), box.icon = "fa-star"}
In order to make the result even more meaningful, we will now make a smoothed map. You can use the `compute_kde` function below.
```

```{r, opts.label = "information"}
library(spatstat)
library(maptools)
library(raster)
compute_kde <- function(feat, adm, title, sigma = 100, res = 50){
  w <- as.owin(as(adm, "Spatial"))
  pts <- st_coordinates(feat)
  p <- ppp(pts[,1], pts[,2], window=w)
  dens <- density.ppp(p, sigma = sigma, eps = res)
  result <- raster(dens, crs = st_crs(adm)[[2]]) * 1000000
  return(result)
}
```

```{block, opts.label = "information"}
Use here a 2km sigma and a 2km resolution and the quantile method (n = 12). 
```

```{r, nm=TRUE, eval=TRUE, echo=FALSE}
library(spatstat)
library(maptools)
library(raster)
compute_kde <- function(feat, adm, title, sigma = 100, res = 50){
  w <- as.owin(as(adm, "Spatial"))
  pts <- st_coordinates(feat)
  p <- ppp(pts[,1], pts[,2], window=w)
  dens <- density.ppp(p, sigma = sigma, eps = res)
  result <- raster(dens, crs = st_crs(adm)[[2]]) * 1000000
  return(result)
}
```


```{r, nm=TRUE, eval=FALSE, solution=TRUE}
x <- compute_kde(feat = sir_31, adm = iris_31, sigma = 2000, res = 2000)
bks <- getBreaks(values(x), nclass = 12, method = "quantile")
cols <- carto.pal("wine.pal", length(bks)-1)
# or 
# cols <- mapview::mapviewGetOption("raster.palette")(length(bks)-1)
par(mar = c(0,0,1.2,0), bg = "#FBEDDA")
plot(st_geometry(iris_31), col = NA, border = NA, main="", bg = "#FBEDDA")
plot(x, breaks = bks, col=cols, add = T,legend=F)
legendChoro(pos = "topleft",
            title.txt = "Restaurants / km2)",values.cex = 0.5,
            breaks = bks, nodata = FALSE,values.rnd = 2,
            col = cols)
layoutLayer(title = "Smoothed Density of Restaurants in Haute-Garonne", scale = 5,
            tabtitle = TRUE, frame = FALSE,
            author = "Kim & Tim", sources = "INSEE - 2018, Sirene - 2018")
```

```{r, sm=TRUE, echo=FALSE}
x <- compute_kde(feat = sir_31, adm = iris_31, sigma = 2000, res = 2000)
bks <- getBreaks(values(x), nclass = 12, method = "quantile")
cols <- carto.pal("wine.pal", length(bks)-1)
# or 
# cols <- mapview::mapviewGetOption("raster.palette")(length(bks)-1)
par(mar = c(0,0,1.2,0), bg = "#FBEDDA")
plot(st_geometry(iris_31), col = NA, border = NA, main="", bg = "#FBEDDA")
plot(x, breaks = bks, col=cols, add = T,legend=F)
legendChoro(pos = "topleft",
            title.txt = "Restaurants / km2)",values.cex = 0.5,
            breaks = bks, nodata = FALSE,values.rnd = 2,
            col = cols)
layoutLayer(title = "Smoothed Density of Restaurants in Haute-Garonne", scale = 5,
            tabtitle = TRUE, frame = FALSE,
            author = "Kim & Tim", sources = "INSEE - 2018, Sirene - 2018")
```

```{block, box.title = "7", box.body = list(fill = "white"), box.icon = "fa-star"}
Try to improve the map resolution.
```

```{r, nm=TRUE, eval=FALSE, solution=TRUE}
x <- compute_kde(feat = sir_31, adm = iris_31, sigma = 2000, res = 500)
bks <- getBreaks(values(x), nclass = 12, method = "quantile")
cols <- carto.pal("wine.pal", length(bks)-1)
# or 
# cols <- mapview::mapviewGetOption("raster.palette")(length(bks)-1)
par(mar = c(0,0,1.2,0), bg = "#FBEDDA")
plot(st_geometry(iris_31), col = NA, border = NA, main="", bg = "#FBEDDA")
plot(x, breaks = bks, col=cols, add = T,legend=F)
legendChoro(pos = "topleft",
            title.txt = "Restaurant Density\n(rest./km2)",values.cex = 0.5,
            breaks = bks, nodata = FALSE,values.rnd = 2,
            col = cols)
layoutLayer(title = "Smoothed Density of Restaurants in Haute-Garonne", scale = 5,
            tabtitle = TRUE, frame = FALSE,
            author = "Kim & Tim", sources = "INSEE - 2018, Sirene - 2018")
```

```{r, sm=TRUE, echo=FALSE}
x <- compute_kde(feat = sir_31, adm = iris_31, sigma = 2000, res = 500)
bks <- getBreaks(values(x), nclass = 12, method = "quantile")
cols <- carto.pal("wine.pal", length(bks)-1)
# or 
# cols <- mapview::mapviewGetOption("raster.palette")(length(bks)-1)
par(mar = c(0,0,1.2,0), bg = "#FBEDDA")
plot(st_geometry(iris_31), col = NA, border = NA, main="", bg = "#FBEDDA")
plot(x, breaks = bks, col=cols, add = T,legend=F)
legendChoro(pos = "topleft",
            title.txt = "Restaurant Density\n(rest./km2)",values.cex = 0.5,
            breaks = bks, nodata = FALSE,values.rnd = 2,
            col = cols)
layoutLayer(title = "Smoothed Density of Restaurants in Haute-Garonne", scale = 5,
            tabtitle = TRUE, frame = FALSE,
            author = "Kim & Tim", sources = "INSEE - 2018, Sirene - 2018")
```

```{block, box.title = "8", box.body = list(fill = "white"), box.icon = "fa-star"}
Would we have obtained the same smoothing for this department if we had kept the restaurants of the bordering departments? Why?
```


```{block, solution = TRUE}
No. The borders would have been different. 
```


# Exercise 4 : Cartogram
```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Make a cartogram of department 31 at the level of intercommunalities in proportion to the number of restaurants (SIRENE data).
```


```{block, opts.label = "information"}
Load the layer ‘epci_31.rds’ from previous exercises and the `cartogram` package. 
```


```{r, nm=TRUE, eval=FALSE, solution = TRUE}
epci_31 <- readRDS("data/epci_31.rds")

library(cartogram)
library(sf)

cartogramme <- cartogram_cont(epci_31, "nb_of_rest", itermax = 5, maxSizeError = 1)
plot(st_geometry(cartogramme), col="lightblue")
```

#### {.tabset}

##### Normal borders
```{r, nm=TRUE, echo=FALSE}
epci_31 <- readRDS("data/epci_31.rds")

library(sf)
plot(st_geometry(epci_31), col="lightblue")
```

##### Cartogram
```{r, nm=TRUE, echo=FALSE}
library(cartogram)
cartogramme <- cartogram_cont(epci_31, "nb_of_rest", itermax = 5, maxSizeError = 1)
plot(st_geometry(cartogramme), col="lightblue")
```

#### {}


# Exercise 5 : Linemap
```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Make a linemap of department 31 at the level of municipalities in proportion to the number of restaurants (SIRENE data).
```

```{block, opts.label = "information"}
Load the map layer ‘com_31.rds’ from previous exercises and the `linemap` package. 
```

```{block, opts.label = "clues"}
Use the two `getgrid` and` linemap` functions of this package.
The following parameters work: `cellsize = 1750`, `k = 400` and `threshold = 0.01`.
```


```{r, nm=TRUE, eval=FALSE, solution = TRUE}
com_31 <- readRDS("./data/com_31.rds")

library(linemap)
grid <- getgrid(x = com_31, cellsize = 1750, var = "nb_of_rest")
plot(sf::st_geometry(com_31), col="ivory1", border = NA)
opar <- par(mar=c(0,0,0,0), bg = "ivory2")
linemap(x = grid, var = "nb_of_rest", k = 400, threshold = 0.01,
        col = "ivory1", border = "ivory4", lwd = 0.6, add = TRUE)
```

```{r, nm=TRUE, echo=FALSE}
com_31 <- readRDS("data/com_31.rds")

library(linemap)
grid <- getgrid(x = com_31, cellsize = 1750, var = "nb_of_rest")
plot(sf::st_geometry(com_31), col="ivory1", border = NA)
opar <- par(mar=c(0,0,0,0), bg = "ivory2")
linemap(x = grid, var = "nb_of_rest", k = 400, threshold = 0.01,
        col = "ivory1", border = "ivory4", lwd = 0.6, add = TRUE)


library(dplyr) #because select is also in raster package and bugs if CACHE = TRUE
```


--------------------------------------------

--------------------------------------------

**reproducibility**

```{r}
sessionInfo()
```
