---
title: "Scraping leboncoin"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r, message=FALSE}
library(rvest)
library(tidyverse)
library(knitr)
```

# Scrape all ads in the real estate category in leboncoin

... for now, just for 1 department: Rhône.

## Get links to all ads

Get the total number of ads and deduce the number of pages to scrape (35 ads are displayed per page).

```{r}
page_base <- "https://www.leboncoin.fr/ventes_immobilieres/offres/rhone_alpes/rhone/"

nb_links <- read_html(page_base) %>% 
  html_nodes("._2ilNG") %>%
  html_text() %>% 
  first() %>% 
  str_replace(" ","") %>% 
  as.numeric()
nb_pages=ceiling(nb_links/35)


pages=c(page_base,
        str_c(page_base,"p-",2:nb_pages))
pages[1:5]
```
Now we have the urls of all the pages we have to scrape to get links to all ads (vector `pages`).

## For each page, get link to individual ads

Definition of function `ads_by_page()` which takes a **page listing ads as an input and returns **all ads' urls** as output.

```{r ads_by_page}
ads_by_page <- function(page){
  my_html <- read_html(
    curl::curl(page,
               handle=curl::new_handle("useragent"="Mozilla/5.0"))
  )
  links <- my_html %>%
      html_nodes(".clearfix") %>% 
      html_attr("href") %>% 
      na.omit()  
  tib <- tibble(urls=str_c("https://www.leboncoin.fr",links)) 
  return(tib)
}  
ads_by_page(pages[1])
```
Now **apply iteratively function `ads_by_page()` to all pages' urls listed in `pages`.

```{r tib_ads_urls}
if(!file.exists("data/tib_ads_urls.csv")){
    tib_ads_urls <- map(pages,safely(ads_by_page)) %>%
      map("result") %>% 
      bind_rows()
    write_csv(tib_ads_urls,"data/tib_ads_urls.csv")
}
tib_ads_urls=read_csv("data/tib_ads_urls.csv")
```
## For each ad, get info

Define function `ad_info()`, which takes **an ad's url** as an input and returns, as an output, a **tibble** with information regarding

- `url`: the ads' urls
- `header`: their headers,
- `date`: the date they were published
- `type`: the type of property
- `surface`: the surface of the property
- `rooms`: the number of rooms
- `GHG`: Greenhouse gas emission category
- `energy_class`: Energy class category,
- `location`: Location of the property

I added some **random waiting time** to each call to `ad_info()` of 0 to 1 second.

```{r ad_info}
ad_info=function(ad){
    Sys.sleep(0.5+runif(1,-0.5,0.5))
    html_ad <- read_html(ad)
    header <- html_ad %>% 
      html_nodes("._1KQme") %>% 
      html_text()
    date <- html_ad %>% 
      html_nodes("._3Pad-") %>% 
      html_text()
    criteria <- 
      tibble(name= html_ad %>% html_nodes("._3-hZF") %>% html_text(),
             value=html_ad %>% html_nodes("._3Jxf3") %>% html_text()) 
    f=function(x){if(length(x)==0){x=NA};return(x)}
    type    <- filter(criteria, name=="Type de bien")$value %>% f()
    surface <- filter(criteria, name=="Surface")$value %>%
      str_extract("^\\d*") %>% f()
    rooms   <- filter(criteria, str_detect(name,"Pi.ces"))$value %>% 
      as.numeric() %>% f()
   
    price <- html_ad %>% 
      html_nodes(".eVLNz") %>% 
      html_text() %>% 
      str_replace_all("[^0-9]","") %>% 
      as.numeric()
    GHG <- html_ad %>% 
      html_nodes("._2BhIP") %>% 
      html_text() %>% 
      first()
    energy_class <-html_ad %>% 
      html_nodes("._15MMC") %>% 
      html_text() %>% 
      first()
    location <- html_ad %>% 
      html_nodes("._1aCZv") %>% 
      html_text() %>% 
      str_replace("Voir sur la carte","")
    tib_ad=bind_cols(urls=ad,
                     header=header,
                     date=date,
                     type=type,
                     surface=surface,
                     rooms=rooms,
                     GHG=GHG,
                     energy_class=energy_class,
                     location=location)
    return(tib_ad)
}
ad_info(tib_ads_urls$urls[1]) %>% kable()
```

Now **apply iteratively this function `ad_info()`** to all ads in `tib_ads_urls`, using `purrr` iteration:

```{r tib_ads}
if(!file.exists("data/tib_ads.csv")){
  tmp=Sys.time()
   tib_ads <- map(tib_ads_urls$urls, safely(ad_info)) %>% 
     map("result") %>% bind_rows()
  Sys.time()-tmp
   write_csv(tib_ads,"data/tib_ads.csv")
}
tib_ads <- read_csv("data/tib_ads.csv")
tib_ads %>% kable()
```

For ~16000 ads, it took about 4 hours for this chunk to run on my computer...


