---
title: "Schelling"
author: "CC&PC"
date: "Summer School"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Schelling Segregation Model

Here are some code lines to prepare the simulation of Schelling's Model of spatial segregation.
We start by the environment.

### Environment

The environment is a static square grid of `size` *  `size` cells, filled at a given percentage `pctOccupied` of householder agents.

A cell can be occupied by a single householder only. 

Each householder belongs to one of two groups, denoted by color (here : Red, Blue).

Basic Schelling model suggest that agents is randomly affected to one groupe or the other (50% of chance), yet we write the code as if it could change in the future.

From the R code point of view, the environment is a matrix, and the state of the system is constituted by the values inside the matrix.
At a given location (i,j), the value of the cell will be either 0 (empty cell), 1 (Red household), or 2 (Blue houshold).

Cells are intialized with 0 at the creation of the `state` matrix.
Households are randomly allocated in the grid by sampling (without replacement) positions.

Household groups are then determined randomly :  the cell value is chosen between 1 and 2 (this time with replacement)  using `pctRed` and `pctBlue`  as probabilities weights.




```{r setupState}


createSchellingState <-  function(size, pctOccupied=.5, pctRed=0.5, pctBlue=1-pctRed ){

state <- matrix(NA,size,size)
nbAgents <- floor(pctOccupied * size^2)
householdCells <- sample(size ^ 2, nbAgents)
state[householdCells] <-  sample(c(1,2), nbAgents, replace = T, prob = c(pctRed,pctBlue))
return(state)
}
state <-  createSchellingState(40,0.5,0.5, 0.5)
```


Now that the sate has been defined, we will code the `display` function, to actually see something.


##Display

Display uses the **ggplot** library, a wierd yet widely used library, with so fancy graphic renderings it seems impossible not to use it.
For the record, we give the non-ggplot version first, using the `image` function. 

N.B. matrix plot is counter-clockwise rotated by an angle of pi/2 ; here we don't care as there is no orientation problems in this theoretic example. 
This is because, by convention, we "naturally" draw *rows* of a matrix along what would be the Y-axis of a plot. 
According to the documentation : 
>  `image` interpret matrix as a table of f(x[i], y[j]) values, so that the x axis corresponds to row number and the y axis to >column number, with column 1 at the bottom, i.e. a 90 degree counter-clockwise rotation of the conventional printed layout of a matrix.

(the same phenomenon occurs with ggplot2 )


```{r displaySimple,eval=FALSE,fig.height=6, fig.width=6}
par(fin=c(6,6)) #plot size setup
image(state, axes = FALSE,col=c("red","blue"))
```


To obtain the ggplot version , the state data structure has to be tweaked a little bit.
ggplot2 needs a data structure where each line is  an individual i.e. a piece of data, formed as follows:
```
    | Var1 | Var2 |    Var3  |
row1| 20   | 12.4 | "Batman" |
row2|  8   | 3.5  | "Robin"  |
```

in our case, pieces of data are householders, located in the grid by their coordinates (`Xcoord,Ycoord`), belonging to a group (`value`); so we need something like: 

```
      | Xcoord | Ycoord | Group |
          ⋮       ⋮       ⋮ 
row_n |   24   |   75   |   2   |
          ⋮       ⋮       ⋮
row_p |   58   |    8   |   1   |
```


Starting from a matrix structure, this transformation is done by the `melt` function.


```{r meltGGplot}
library(reshape2)
meltedState <- melt(state, varnames = c("X","Y")) 
head(meltedState,10)  #to see what it looks like and get the column names
```



```{r displayGGplot, fig.height=6, fig.width=6}
library(ggplot2)
# color definitions  
colorMapping <- c("1"="red", "2"="blue") 

myggplot <- ggplot(meltedState, aes(x=X, y=Y)) +
  geom_raster(aes(fill=factor(value)))+
 scale_fill_manual(values = colorMapping, na.value="lightgrey")+
  theme_void()
myggplot 


```

this does the job, 
yet we can improve rendering by drawing household in cells as points, in a scattermap-like way:


```{r displayGGplotfancy, fig.height=6, fig.width=6}
# color definitions  
library(RColorBrewer)
#fancyColors <- brewer.pal(3,"Accent")
# declaration of colors by name
fancyColors <- c("1"="bisque", "2"="aquamarine") 

myggplot2 <- ggplot(meltedState, aes(x=X, y=Y)) +
  geom_point(size=1, aes(colour=factor(value)))+ #size has to be adjusted regarding the size of the state
 scale_color_manual( values=fancyColors, name="Group" )+
  theme_void()+
  coord_fixed(ratio = 1) + 
  ggtitle("Schelling state")
   
myggplot2            

#save as a function for later use
displayState <- function(mystate, dotSize=1){
 myggplot2 <- ggplot(mystate, aes(x=X, y=Y)) +
  geom_point(size=dotSize, aes(colour=factor(value)))+ #size has to be adjusted regarding the size of the state
 scale_color_manual( values=fancyColors, name="Group" )+
  theme_void()+
  coord_fixed(ratio = 1) + 
  ggtitle("Schelling state")
  return(myggplot2) 
}




```



## Dynamics

Each household "aims"" at finding a location where the proportion of similar households in the neighborhood (surrounding cells) is above a certain `tolerance` threshold. 
If a household is 
   
 At each simulation step, each agent :
 
 1. looks around and establish the proportion of similar householder in its (Moore) nieghborhood
 2. if this proportion is below the threshold → looks for another place where the proportion is above , remains in place otherwise.
 
 
 To achieve this beahavior , we need a way to figure out :

* the neighborhood af a householder
* evaluate householder "happiness" i.e. if the neighborhood fits the tolerance criterion.
* locations of  the empty cells of the grid
* locations of the adequate cells for each group are (i.e. whose neighborhood proportion of same group is below the tolerance threshold)


### Getting the neighborhood

We will start with the Moore neighborhood (the 8 adjacents cells of a cell) ; this defintion could be extended in a later version.


Given a coordinate (i,j) in the state matrix, the neighborhood is given by the following code. 
The `meAndMyNeighbors(i,j)` select the neighboring portion of the (i,j) cell in the `state` matrix.
The `getgetneighbors(i,j)` transforms this portion in a more convenient format (a list) and excludes the calling cell (i,j) 
Special cases of corner and border cells are considered having incomplete neighborhood.

```{r neighbors}
meAndMyNeighbors <- function (state,i,j) {
#we assume world is square for now
    size <-  nrow(state)  
    # corners 
  if (j==1 & i==1 ){
    return(state[c(i, i + 1), c(j, j + 1)])
  } 
  if (j==size & i ==size){
    return(state[c(i-1, i), c(j-1, j)])
  } 
  if(i==size & j==1) {
    return(state[c(i-1, i ), c(j, j + 1)])
  }
  if(i==1 & j==size){
    return(state[c(i, i + 1), c(j-1, j)])
  }
  
  #borders without corners
  if (i==1 &  j > 1 & j < size) {
    return(state[c(i, i + 1), c(j-1,j, j+1)])
  }
  if(i==size & j >1 & j < size ){
    return(state[c(i-1, i), c(j-1,j, j+1)])
  }
  if(j==1 & i >1 & i < size){
        return(state[c(i-1, i, i+1), c(j, j+1)])
  }
  if(j==size & i >1 & i < size){
    return(state[c(i-1, i, i+1), c(j-1, j)])
  }
  #regular cell case
  if (i > 1 & i < size & j > 1 & j < size){
    return(state[c(i-1,i,i+1),c(j-1,j,j+1)])
  }
}

getneighbors <- function(state, i,j){
    neighborhood <- meAndMyNeighbors(state, i, j)
        # flatten the 3xn matrix as a list
    neighborhood <- as.vector(neighborhood)
  return(neighborhood)
}



```



### Evaluating cells "happiness"

Now we need to evaluate if the householder of a cell (i,j) is satisfied by its neighborhood , i.e. if the proportion of similar householders around him is  above the `tolerance` threshold.


````{r happiness}
tolerance <- 0.3  


isHappy <- function(state, i , j, tolerance){
  cell <-  state[i,j]
  if (is.na(cell)){return(NA)}
  neigh <- getneighbors(state,i,j)
  numberOfSimilar <-sum(cell==neigh, na.rm = T)  
  happy <-  numberOfSimilar / (length(neigh) ) > tolerance
  return(happy)  
} 
```

To test the function, we have to apply it to each cell of the matrix.
This will be more convenient if we use the melted state, where coordinates are immediately available for each cell.

``` {r testhappiness}
library(dplyr)

testEnv <-  createSchellingState(10,pctOccupied = 0.8)



# we use melted state for immediatley accessible coordinates
meltTestEnv <-  melt(testEnv,varnames = c("X","Y"))
displayState(meltTestEnv, dotSize = 5)
#handy rowwise operator form dplyr
meltTestEnv <-  meltTestEnv %>% rowwise() %>% mutate(happy= isHappy(state = testEnv,X,Y,0.3))

# we update the display function code
displayStateUnhappy <- function(meltedState){
  pp <- displayState(meltedState, dotSize = 5) + 
    geom_text(data = subset(meltedState, !happy ), aes(X,Y,label = ":-|"), angle=-90) 
return(pp)
  }



# save as function for later
getUnhappy <-  function(meltedState){
  return(subset(meltedState, !happy))
}

createMeltedSchellingState <- function(size, pctOccupied=.5, pctRed=0.5, pctBlue=1-pctRed, tolerance){
  matState <-  createSchellingState(size, pctOccupied, pctRed, pctBlue)
  meltedState <-  melt(matState,varnames = c("X","Y"))
  meltedState <-  meltedState %>% rowwise() %>% mutate(happy= isHappy(state = matState,X,Y, tolerance ))
  # we add an explicit ID of cells (row number), it will be useful
  meltedState$ID <- seq.int(nrow(meltedState))
  return(meltedState)
}




``` 



### Getting empty cells

Unsatisfied householders look for empty cells to relocate.
In our model, empty cells have a `NA`value.
We will create and maintain a subset of our meltedState, keeping track of the empty cells.
We use a function that filter the values of the melted state.

```{r }
getEmptyCells <-  function(meltedState){
  return(meltedState %>%filter(is.na(value))  )
}
emptyCells <-  getEmptyCells(meltTestEnv)

#we add a symbol layer  to check emptyness visually
pp <-  displayStateUnhappy(meltTestEnv) +
  geom_text(data = subset(meltTestEnv, is.na(value) ), aes(X,Y,label = sprintf("\U2205")), angle=-90)  
pp


#we update the function 

displayStateUnhappy <- function(mystate) {
  pp <- displayState(mystate , dotSize = 5) +
  geom_text(data = subset(mystate,!happy),  aes(X, Y, label = ":-|"),   angle = -90) +
  
  geom_text(  data = subset(mystate, is.na(value)),   aes(X, Y, label = sprintf("\U2205")),   angle = -90)
  
  return(pp)
}


```


### Moving to more satisfying locations

The idea is now to move unhappy households to another, empty location . 
For each householder, an empty location/cell is randomly drawn, and the householder establish itself in it .

To do that, we sample among empty cells, update these empty cells values according to unhappy householders values, and put the previous unhappy householders cells to`NA` (i.e. empty)



```{r move}

stateEnv <- createMeltedSchellingState(10,pctOccupied = 0.8, tolerance=0.3)

unhappy <-  getUnhappy(stateEnv)
numberOfUnHappy <-  nrow(unhappy)
emptyCells <-  getEmptyCells(stateEnv)
numberOfEmpty <-  nrow(getEmptyCells(stateEnv))
cat(numberOfEmpty , "\n") # has to be 20% since pctOccupied is 0.8 of size squared


# if there is more empty room than unhappy, we can move everyone in a single loop 
if (numberOfEmpty > numberOfUnHappy){

    futureOccupiedCells <-  sample_n(emptyCells, numberOfUnHappy)
    
    #when household settle , values are updated
    futureOccupiedCells$value <-  unhappy$value
    
    #newly occupied cells values are no longer NA , but the values of unhappyhouseholders
    stateEnv[futureOccupiedCells$ID,"value"] <-  futureOccupiedCells$value
    #previously occupied cells are set to empty
    stateEnv[unhappy$ID, "value"] <- NA 
}

# if there is more unhappy than empty cells, we can only move a subset of unhappy householders to empty locations
#code looks similar, notice the replacement of unhappy by a subset : futureMovers
if (numberOfEmpty < numberOfUnHappy){
  
    futureMovers <-  sample_n(unhappy, numberOfEmpty)
     futureOccupiedCells <-  emptyCells
    #when household settle , values are updated
    futureOccupiedCells$value <-  futureMovers$value
                                           
                                           
    #newly occupied cells values are no longer NA , but the values of unhappyhouseholders
    stateEnv[futureOccupiedCells$ID,"value"] <-  futureOccupiedCells$value
    #previously occupied cells are set to empty
    stateEnv[futureMovers$ID, "value"] <- NA 
  }


```


## Simulation Loop

Now that we have implemented one step of the model Dynamic , we can automate its execution in a loop, create some buttons and animate it in a `shiny`app (its a specific library that conveniently handles Graphical User Interface stuff)

Before we do that, we add the last lines of code in a function, 

```{r loopsimu}

step <- function(meltedState){
  unhappy <-  getUnhappy(meltedState)
  numberOfUnHappy <-  nrow(unhappy)
  emptyCells <-  getEmptyCells(meltedState)
  numberOfEmpty <-  nrow(getEmptyCells(meltedState))
if (numberOfEmpty >= numberOfUnHappy){
    futureOccupiedCells <-  sample_n(emptyCells, numberOfUnHappy)
    futureOccupiedCells$value <-  unhappy$value
    meltedState[futureOccupiedCells$ID,"value"] <-  futureOccupiedCells$value
    meltedState[unhappy$ID, "value"] <- NA 
}

if (numberOfEmpty < numberOfUnHappy){
    futureMovers <-  sample_n(unhappy, numberOfEmpty)
     futureOccupiedCells <-  emptyCells
    futureOccupiedCells$value <-  futureMovers$value
    meltedState[futureOccupiedCells$ID,"value"] <-  futureOccupiedCells$value
    meltedState[futureMovers$ID, "value"] <- NA 
}
  return(meltedState)
}


simulate <-  function(steps, state){
  for (i in 1:steps){
    state <-  step(state)
    
  }
}




```



Basically a simulation loop is a loop that repeat a simulation step a certain number of times or until a certain condition is reached.


```{r, echo=FALSE}
library(shiny)
shinyApp(
  ui = fluidPage(
     actionButton("go", "Go"),
   plotOutput("plot")
    
  ),

  server = function(input, output) {
  
    state <-createMeltedSchellingState(20, 0.8, 0.5, 0.5,0.3)  
    
  currentState <- eventReactive(input$go, {
    step(state)
  })

  output$plot <- renderPlot({
    currentplot <-  displayStateUnhappy(currentState())
    currentplot
  })  }
  ,
   options = list(height = 500)
 )
```




