---
title: "Schelling"
author: "CC&PC"
date: "Summer School"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Schelling Segregation Model

Here are some code lines to prepare the simulation of Schelling's Model of spatial segregation.
We start by the environment.

### Environment

From now on , the environment is a static square grid of `size` *  `size` cells, filled at a given percentage `pctOccupied` of householder agents.

A cell can contain only one householder. 

Each householder belongs to one of two groups (for now), denoted by values (here : 1 and 2).

Basic Schelling model suggest that agents is randomly affected to one groupe or the other (50% of chance), yet we write the code as if it could change.

From the R code point of view, the environment grid is easier to consider as a matrix, even if it will be soon transformed in a long format data table, moreconvenient for `dplyr` library functions.


At a given location (i,j), the value of the cell will be either `NA` (empty cell), 1 (Red household), or 2 (Blue houshold).

At initialization,  households are randomly located in the grid by sampling (without replacement) positions.

Household groups are then determined randomly :  the cell value is chosen between 1 and 2 (this time with replacement)  using `pctRed` and `pctBlue`  as probabilities weights.




```{r setupState}


matrixSchellingState <-  function(size, pctOccupied=.5, pctRed=0.5, pctBlue=1-pctRed ){

matrixState <- matrix(NA,size,size)
nbAgents <- floor(pctOccupied * size^2)
householdCells <- sample(size ^ 2, nbAgents)
matrixState[householdCells] <-  sample(c(1,2), nbAgents, replace = T, prob = c(pctRed,pctBlue))
return(matrixState)
}
matrixState <-  matrixSchellingState(40,0.5,0.5, 0.5)
```


Now that the sate has been defined, we will code the `display` function, to actually see something.


##Display

### Data Format

Display uses the **ggplot** library, a widely used library, with so fancy graphic renderings it seems impossible not to use it.

To that extent, the state data structure has to be tweaked a little bit.
ggplot2 needs a data structure where each line is  an individual i.e. a piece of data, formed as follows:
```
     | Var1 | Var2 |    Var3  |
|----|------|------|----------|
|row1| 20   | 12.4 | "Batman" |
|row2|  8   | 3.5  | "Robin"  |
```

in our case, pieces of data are householders, located in the grid by their coordinates (`Xcoord,Ycoord`), belonging to a group (`value`); so we need something like: 

```
      | Xcoord | Ycoord | Group |
          ⋮       ⋮       ⋮ 
row_n |   24   |   75   |   2   |
          ⋮       ⋮       ⋮
row_p |   58   |    8   |   1   |
```


Starting from a matrix structure, this transformation is done by the `melt` function of `reshape` library.


```{r meltGGplot}
library(reshape2)
meltedState <- melt(matrixState, varnames = c("X","Y")) 
head(meltedState,10)  #to see what it looks like and get the column names
```

### Display as raster


```{r displayGGplot, fig.height=6, fig.width=6, warning =FALSE}
library(ggplot2)
# color definitions  
colorMapping <- c("1"="red", "2"="blue") 

myggplot <- ggplot(meltedState, aes(x=X, y=Y)) +
  geom_raster(aes(fill=factor(value)))+
 scale_fill_manual(values = colorMapping, na.value="lightgrey")+
  theme_void()
myggplot 


```

this does the job, 
yet we can improve rendering by drawing household in cells as points, in a scattermap-like way:

### Display as scatterplot
`ggplot` will whine because of the NA values of empty cells, you can ignore the warnings.


```{r displayGGplotfancy, fig.height=6, fig.width=6}
# color definitions  
fancyColors <- c("1"="bisque", "2"="aquamarine") 

myggplot2 <- ggplot(meltedState, aes(x=X, y=Y)) +
  geom_point(size=2, aes(colour=factor(value)))+ #size has to be adjusted regarding the size of the state
 scale_color_manual( values=fancyColors, name="Group" )+
  theme_void()+
  coord_fixed(ratio = 1) + 
  ggtitle("Schelling state")
   
myggplot2            

#save as a function for later use
displayState <- function(mystate, dotSize=1){
 myggplot2 <- ggplot(mystate, aes(x=X, y=Y)) +
  geom_point(size=dotSize, aes(colour=factor(value)))+ #size has to be adjusted regarding the size of the state
 scale_color_manual( values=fancyColors, name="Group" )+
  theme_void()+
  coord_fixed(ratio = 1) + 
  ggtitle("Schelling state")
  return(myggplot2) 
}
```



## Dynamics




Each household "aims"" at finding a location where the proportion of similar households in the neighborhood (surrounding cells) is above a certain `tolerance` threshold. 
If a household is 
   
 At each simulation step, each agent :
 
 1. looks around and establish the proportion of similar householder in its (Moore) nieghborhood
 2. if this proportion is below the threshold → looks for another place where the proportion is above , remains in place otherwise.
 
 
 To achieve this beahavior , we need a way to figure out :

* the neighborhood af a householder
* evaluate householder "happiness" i.e. if the neighborhood fits the tolerance criterion.
* locations of  the empty cells of the grid
* locations of the adequate cells for each group are (i.e. whose neighborhood proportion of same group is below the tolerance threshold)


### Getting the neighborhood

We will start with the Moore neighborhood (the 8 adjacents cells of a cell) ; this defintion could be extended in a later version.


Given a coordinate (i,j) the neighborhood is given by the following code. 
The `meAndMyNeighbors(i,j)` select the neighboring portion of the (i,j) cell in the `state` .
The `getgetneighbors(i,j)` transforms this portion in a more convenient format (a list) and excludes the calling cell (i,j) 
Special cases of corner and border cells are considered having incomplete neighborhood.

```{r neighbors}
library(dplyr)
#to ensure the call to filter function is calling the dplyr version
filter = dplyr::filter
#same function but for a state in the long format
meAndMyNeighbors <- function (state,i,j) {
  #this version doesn't suppose a square world, but a 9 Moore neighborhod
  neigh <-   filter(state, between(X,i-1,i+1) & between(Y,j-1,j+1) & !is.na(value))
  return(neigh)
}


getneighbors <- function(state, i,j){
    neighborhood <- meAndMyNeighbors(state, i, j)
    #exclude the caller cell
    neighborhood <- neighborhood[!(neighborhood$X==i & neighborhood$Y==j),]
  return(neighborhood)
}
```



### Evaluating cells "happiness"

Now we need to evaluate if the householder of a cell (i,j) is satisfied by its neighborhood , i.e. if the proportion of householders of another group  around him is  below the `tolerance` threshold.


````{r happiness}
tolerance <- 0.3  

isHappy <- function(state, i , j, tolerance){
  cell <-  filter(state, X==i & Y==j)
  if (is.na(cell$value)){return(NA)}
  neigh <- getneighbors(state,i,j) 
  numberOfDiff <-sum(cell$value!=neigh$value, na.rm = T)  
  happy <-  (numberOfDiff / nrow(neigh) ) < tolerance
  return(happy)  
} 
```

We now create a test environment and apply the function `isHappy` to each cell.

``` {r testhappiness}


testEnv <-  matrixSchellingState(5,pctOccupied = 0.8)
meltTestEnv <-  melt(testEnv,varnames = c("X","Y"))
displayState(meltTestEnv, dotSize = 5)



#we apply the function isHappy to create a newcolumn of happiness in the state
meltTestEnv <- meltTestEnv %>% rowwise() %>% mutate(happy= isHappy(., X, Y, tolerance = 0.3))


# we update the display function code to display unhappiness
displayStateUnhappy <- function(meltedState){
  pp <- displayState(meltedState, dotSize = 5) + 
    geom_text(data = subset(meltedState, !happy ), aes(X,Y,label = ":-|"), angle=-90) 
return(pp)
  }
displayStateUnhappy(meltTestEnv)




# save as function for later
updateHappiness <- function(state, tolerance){
  state <- state %>% rowwise() %>% mutate(happy= isHappy(., X, Y, tolerance ))
}

getUnhappy <-  function(state){
  return(filter(state, !happy))
}
#melted state constructor
createMeltedSchellingState <- function(size, pctOccupied=.5, pctRed=0.5, pctBlue=1-pctRed, tolerance){
  matState <-  matrixSchellingState(size, pctOccupied, pctRed, pctBlue)
  # Warning : melt destroy attr
  meltedState <-  melt(matState,varnames = c("X","Y"))
  meltedState <-  updateHappiness(meltedState, tolerance  )
  # we add an explicit ID of cells (row number), it will be useful
  meltedState$ID <- seq.int(nrow(meltedState))
  return(meltedState)
}
``` 



### Getting empty cells

Unsatisfied householders look for empty cells to relocate to.
In our model, empty cells have a `NA`value.
We create and maintain a subset of our meltedState, keeping track of the empty cells.


```{r }
getEmptyCells <-  function(state){
  return(state %>%filter(is.na(value))  )
}
emptyCells <-  getEmptyCells(meltTestEnv)

#we add a symbol layer  to check emptyness correction  visually
pp <-  displayStateUnhappy(meltTestEnv) +
  geom_text(data = subset(meltTestEnv, is.na(value) ), aes(X,Y,label = sprintf("\U2205")), angle=-90)  
pp


```


### Moving to more satisfying locations

The idea is now to move unhappy households to another, empty location . 
For each householder, an empty location/cell is randomly drawn, and the householder moves in it .

We sample among empty cells, update these empty cells values according to unhappy householders values, and put the previous unhappy householders cells to`NA` (i.e. empty)

We start by moving householders one at a time, via the `moveOne` function.

```{r moveOne}
#we create another test environemment and get rid of the previous
rm(meltTestEnv)
stateEnv <- createMeltedSchellingState(10,pctOccupied = 0.8, tolerance=0.3)

displayStateUnhappy(stateEnv)
unhappy <-  getUnhappy(stateEnv)
numberOfUnHappy <-  nrow(unhappy)
emptyCells <-  getEmptyCells(stateEnv)
numberOfEmpty <-  nrow(emptyCells)
cat(numberOfEmpty , "\n") # has to be 20% of state size squared since pctOccupied is 80%

moveOne <-  function(state, tolerance, followMover=F){
  #take one unhappy householder 
  unhappyHHs <-  getUnhappy(state)
  uhhh <- sample_n(unhappyHHs, 1)
  emptyCells <-  getEmptyCells(state)
  emptyCell <-  sample_n(emptyCells,1)
  
  if (followMover){
  pp <-  displayStateUnhappy(state)+
    geom_point(data = filter(state, ID==uhhh$ID ), aes(X,Y),color="red", size = 8)+
    geom_point(data = filter(state, ID==emptyCell$ID ), aes(X,Y),color="green", size = 8)

  print(pp)
}
  
  #State has to be updated by swapping values of these two cells , ID is useful here
   state[state$ID==emptyCell$ID,"value"] <- uhhh$value
  state[state$ID==uhhh$ID,"value"] <- emptyCell$value

  #moving changes happiness, so we update the state
  state <-  updateHappiness(state, tolerance )

    return(state)
}
```


To check the dynamic , we consider 3 plots. 
First one is state before we move one householder.
Next plot show the sampled householder with a red circle, and its destination(empty cell) with a green one.
Moving happens, and the final plot show the state afterwards

```{r moveOneSequence}
displayStateUnhappy(stateEnv)
stateEnv <- moveOne(stateEnv, 0.3, followMover = T)
displayStateUnhappy(stateEnv)
```

Notice that moving to a cell may update happiness around it, and around the previous householder location as well. 




## Simulation Loop

Now that we have implemented one step of the model Dynamic , we can automate its execution in a loop, create some buttons and animate it in a `shiny` app (a library for GUI and reactive programming stuff)


```{r loopsimu}
step <- function(state, tolerance){
  unhappy <-  getUnhappy(state, tolerance)
  if (nrow(unhappy) == 0){
    return(state)
  }
    state <- moveOne(state,tolerance)
  return(state)
}


simulate <-  function(steps, state, tolerance){
  for (i in 1:steps){
    state <-  step(state, tolerance)
  }
}
```



Basically a simulation loop is a loop that repeat a simulation step a certain number of times or until a certain condition is reached. Here we chose to execute simulation steps a certain number of times (and do nothing in case everyone is happy)


```{r, echo=FALSE}
library(shiny)
shinyApp(
  ui = fluidPage(
     actionButton("go", "Move one"),
   plotOutput("plot")
  ),

  server = function(input, output) {
    tolerance <-  0.3
    state <-createMeltedSchellingState(20, 0.8, 0.5, 0.5,tolerance)  
    currentState   <- eventReactive(input$go, {
    state <-  step(state, tolerance)
  })

  output$plot <- renderPlot({
    currentplot <-  displayStateUnhappy(currentState())
    currentplot
  })  }
  ,
   options = list(height = 500)
 )
```



# Measures

## Global Similarity percentage

Global similarity is the aggregated (or averaged) ratio between similar neighbors and neighbors. On perfect segregation, it should be of 100%.

```{r globalSimilarity }
getSimilarNeighborsNumber <-  function(state, i, j){
  #cell <-  filter(state, X==i & Y==j)
  cell <-  state[state$X==i & state$Y==j, ]
  neigh <- getneighbors(state,i,j) 
  numberOfSimilar <-sum(cell$value==neigh$value, na.rm = T)  
return(numberOfSimilar)
    }
  
 
globalSimilarityPct <-  function(state){
  #res <-  state %>% rowwise() %>% transmute(ratioSim = nrow(getneighbors(.,X,Y)) / nrow(getSimilarNeighbors(.,X,Y)))
  res <-  state %>% rowwise() %>% transmute(nbSimNeigh = getSimilarNeighborsNumber(.,X,Y) , nbNeigh = nrow(getneighbors(.,X,Y)))

    return(sum(res$nbSimNeigh) / sum(res$nbNeigh))
}
globalSimilarityPct(stateEnv)
```



## Global unsatisfaction percentage 


Let's define roughly satisfaction as "not being unhappy", which can be evaluated by counting the number of householders that do not want to move, normalized by the total population. In a stable configuration, it should be 100%


```{r globalSatisfaction }
globalunHappyPct <-  function(state){
  populationSize  <-  nrow(state %>%  filter(!is.na(value)))
  sads <- nrow(state %>%  filter(!happy))
    return(sads/populationSize * 100)
}
globalunHappyPct(stateEnv)
```



## Segregation measures




# Alterations 

Since schelling's segregation model is a simple model , a lot of *alterations* can be made :

* Choice of location criteria
* happiness evaluation
* neighborhood sizes
* number of groups 
* emptiness of cells
* radius threshold when moving
* rates of moving


Some of them can be madfe by changing parameters values, most of them requires some new code writing.


##Exodus, movement of Jah people

For now, we have moved householder one at a time. 
The following code shows a way of moving as many householders as possible.

Two cases : if there is enough enough empty cells to move unhappy householder , we sample in empty cells the adequate number of locations, otherwise , we sample in unhappy householders the number of empty cells.

```{r moveSeveral}



moveSeveral <- function(state, tolerance) {
  uHHHs <-  getUnhappy(state)
  emptyCells <-  getEmptyCells(state)
  numberOfEmpty <-  nrow(getEmptyCells(state))
  numberOfUnHappy <-  nrow(getUnhappy(state))
  
  # if there is more empty room than unhappy, we can move everyone in a single move
    if (numberOfEmpty > numberOfUnHappy) {
  futureOccupiedCells <-  sample_n(emptyCells, numberOfUnHappy)
  
  #State has to be updated by swapping values of these two cells , ID is useful here
  state[state$ID == emptyCells$ID, "value"] <- uHHHs$value
  state[state$ID == uHHHs$ID, "value"] <- emptyCells$value
  
  #moving changes happiness, so we update the state
  state <-  updateHappiness(state, tolerance)
  
  }
  
  # if there is more unhappy than empty cells, we subset householders as futureMovers
  if (numberOfEmpty < numberOfUnHappy) {
  futureMovers <-  sample_n(unhappy, numberOfEmpty)
  futureOccupiedCells <-  emptyCells
  
#State has to be updated by swapping values of these two cells , ID is useful here
  state[state$ID == emptyCells$ID, "value"] <- futureMovers$value
  state[state$ID == futureMovers$ID, "value"] <- emptyCells$value
  
  #moving changes happiness, so we update the state
  state <-  updateHappiness(state, tolerance)
  }
  return(state)
}

```

