---
title: "Schelling"
author: "PC&CC"
date: "Summer School"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Schelling Segragation Model

Here are some code lines to prepare the simulation of Schelling's Model of spatial segregation.
We start by the environment.

### Environment

The environment is a static square grid of `size` *  `size` cells, filled at a given percentage `pctOccupied` of householder agents.
A cell can be occupied by a single householder only. 
Each householder belongs to one of two groups, denoted by color (here : Red, Blue).
Basic Schelling model suggest that agents is randomly affected to one groupe or the other (50% of chance), yet we write the code as if it could change in the future.

From the R code point of view, the environment is a matrix, and the state of the system is constituted by the values inside the matrix.
At a given location (i,j), the value of the cell will be either 0 (empty cell), 1 (Red household), or 2 (Blue houshold).

Cells are intialized with 0 at the creation of the `state` matrix.
Households are randomly allocated in the grid by sampling (without replacement) positions.

Household groups are then determined randomly :  the cell value is chosen between 1 and 2 (this time with replacement)  using `pctRed` and `pctBlue`  as probabilities weights.




```{r setupState}
size <- 100
pctOccupied <- 0.5
pctRed <- 0.5
pctBlue <- (1-pctRed)

state <- matrix(NA,size,size)
nbAgents <- floor(pctOccupied * size^2)
householdCells <- sample(size ^ 2, nbAgents)
state[householdCells] <-  sample(c(1,2), nbAgents, replace = T, prob = c(pctRed,pctBlue))
```


Now that the sate has been defined, we will code the `display` function, to actually see something.


##Display

Display is achived by the **ggplot** library, a kind of wierd and fucked-up piece of software, but featuring so fancy graphic renderings,  you'll be seen as an old twat if you don't keep up and use it.

For the record, we give the non-ggplot version first, using the `image` function. 
N.B. matrix plot is counter-clockwise rotated by an angle of pi/2 ; here we don't care as there is no orientation problems in this theoretic example. 
```{r displaySimple,fig.height=6, fig.width=6}
par(fin=c(6,6))
image(state, axes = FALSE,col=c("red","blue"))
```

This is beacause, by convention, we "naturally" draw *rows* of a matrix along what would be the Y-axis of a plot. 
According to the documentation : 

> `image` interpret matrix as a table of f(x[i], y[j]) values, so that the x axis corresponds to row number and the y axis to >column number, with column 1 at the bottom, i.e. a 90 degree counter-clockwise rotation of the conventional printed layout of a matrix.

(the same phenomenon occurs with ggplot2 )


To obtain the ggplot version , the state data structure has to be tweaked a little bit.
ggplot2 needs a datastructure where each line is formed as follows:
```
Xcoord | Ycoord | value
```

Starting from a matrix structure, this transformation is done by the `melt` function.


```{r displayGGplot, fig.height=6, fig.width=6}
library(ggplot2)
library(reshape2)
meltedState <- melt(state) 
head(meltedState,10)  #to see what it looks like and get the column names

# color definitions  
colorMapping <- c("1"="red", "2"="blue") 


myggplot <- ggplot(meltedState, aes(x=Var1, y=Var2)) +
  geom_raster(aes(fill=factor(value)))+
 scale_fill_manual(values = colorMapping)+
  theme_void()
myggplot            

```


## Dynamics

Each household "aims"" at finding a location where the proportion of similar households in the neighborhood (surrounding cells) is above a certain `tolerance` threshold.
   
 At each simulation step, each agent :
 
 1. looks around him (Moore nieghborhood)
 2.
 
  
 

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
