---
title: "Modelling"
author: "CC&PC"
date: "Summer School"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

C. Anderson provokingly in the magazine Wired: 

> "The End of Theory: the data deluge makes the scientific method obsolete""

## Why model?


J. Epstein (2008, JASSS): 'Why model?'

* Explain (very distinct from predict)
* Guide data collection
* Illuminate core dynamics
* Suggest dynamical analogies
* Discover new questions
* Promote a scientific habit of mind
* Bound (bracket) outcomes to plausible ranges
* Illuminate core uncertainties.
* Offer crisis options in near-real time
* Demonstrate tradeoffs / suggest efficiencies
* Challenge the robustness of prevailing theory through perturbations
* Expose prevailing wisdom as incompatible with available data
* Train practitioners
* Discipline the policy dialogue
* Educate the general public
* Reveal the apparently simple (complex) to be complex (simple)

A model is a simplified representation/abstraction of a target system, which implements some theoretical propositions about the logical linkages between objects of interest.

In this tutorial, two kinds of models are presented: 

* descriptive models (statistics)
* genrative models (ABM)

More generally, we want to show how the two shed complementary lights on spatial problems and how they interact with the new massive data.

### Case study: urban segregation

Let's use the information theory to qualify *diversity* and *segregation* of a given city (cf. John Iceland et al on multigroup entropy: https://www.census.gov/hhes/www/housing/resseg/multigroup_entropy.pdf). The measure have been implemented are described below and tested on Canadian metropolisis from package `cancensus` (example by @dshkol: https://github.com/dshkol/scratchpad/blob/master/content/post/2018-05-10-diversity-and-segregation-i.Rmd).


#### Diversity
```{r diversity}

diversity_index <- function(cma) {
  
  cma.ct <- get_census("CA16", regions=list(CMA=cma), 
                       vectors = minority_vectors, level = "CT",
                       labels = "short", geo_format = NA)
  
  cma.csd <- get_census("CA16", regions=list(CMA=cma), 
                        vectors = minority_vectors, level = "CSD",
                        labels = "short", geo_format = NA)
  
  # Calculating diversity (Theil's E)
  # For every variable, divide by v_CA16_3999 and multiply times the logged inverse proportion, then
  # take the sum for each tract. With 14 different groups, the max entropy is ln(14) = 2.64
  base_pop <- quo(v_CA16_3954)
  cma.ei <- cma.ct %>% 
    group_by(GeoUID) %>% 
    mutate_at(minorities, funs(E = (./!!(base_pop))*(log(!!(base_pop)/.)))) %>%
    select(GeoUID, ends_with("_E")) %>% 
    ungroup() %>% 
    mutate_at(vars(ends_with("_E")), funs(ifelse(is.nan(.),0,.))) %>% 
    mutate(Ei = rowSums(select(.,-1), na.rm = FALSE)) %>% 
    select(GeoUID, Ei)
  
  cma.eicsd <- cma.csd %>% 
    group_by(GeoUID) %>% 
    mutate_at(minorities, funs(E = (./!!(base_pop))*(log(!!(base_pop)/.)))) %>%
    select(GeoUID, ends_with("_E")) %>% 
    ungroup() %>% 
    mutate_at(vars(ends_with("_E")), funs(ifelse(is.nan(.),0,.))) %>% 
    mutate(Ei = rowSums(select(.,-1), na.rm = FALSE)) %>% 
    select(GeoUID, Ei)
  
  # Join with geography
  
  cma.geo <- get_census_geometry("CA16", regions=list(CMA=cma), 
                                 level = "CT", geo_format = "sf")
  
  cma.csd.geo <- get_census_geometry("CA16", regions=list(CMA=cma), 
                                     level = "CSD", geo_format = "sf")
  
  # Which Census Tracts are in which CSD
  
  cma.ct <- cma.geo %>% 
    left_join(cma.ei) %>% 
    mutate()
  
  cma.csd <- cma.csd.geo %>% 
    left_join(cma.eicsd) %>% 
    mutate()
  
  # Adding map detail
  mz_set_tile_host_nextzen()
  get_vector_tiles <- function(bbox){
    mz_set_tile_host_nextzen(getOption("nextzen_API_key"))
    mx_box=mz_rect(bbox$xmin,bbox$ymin,bbox$xmax,bbox$ymax)
    mz_vector_tiles(mx_box, height = 1000, width = 1000)
  }
  
  bbox <- st_bbox(cma.ct)
  vector_tiles <- get_vector_tiles(bbox)
  
  if(length(vector_tiles$water$features) > 0) {
    water <- as_sf(vector_tiles$water)
  } 
  
  if(length(vector_tiles$roads$features) > 0) {
    roads <- as_sf(vector_tiles$roads) %>% 
      filter(kind == "highway")
  } else {roads <- water}
  
  if(length(vector_tiles$transit$features) > 0) {
    transit <- as_sf(vector_tiles$transit) %>% filter(kind == "subway")
  } else {transit <- water}
  
  ct_div_plot <- ggplot(cma.ct) + 
    geom_sf(aes(fill = Ei, colour = Ei)) +  
    geom_sf(data = cma.csd.geo, fill = NA, colour = "white") +
    geom_sf(data = transit, size = 0.2, colour = "grey24") +
    geom_sf(data = roads, size = 0.2, colour = "grey36") +
    geom_sf(data = water, fill = "lightblue", colour = NA) + 
    coord_sf(datum = NA) +
    scale_fill_viridis_c("Diversity Entropy Index",
                         option = 3, breaks = c(0,0.5,1,1.5,2), 
                         limits = c(0,2),
                         labels = c("Less\nDiverse","","","","More\nDiverse"),
                         guide = guide_legend(
                           direction = "horizontal",
                           title.position = "top",
                           label.position = "bottom",
                           keywidth = unit(2,"line"))) +
    scale_colour_viridis_c(option = 3, guide = "none",limits = c(0,2)) +
    theme(panel.background = element_blank(),
          legend.position = c(0.2,0.9),
          legend.background = element_blank(),
          legend.key = element_rect(color = NA)) +
    labs(caption = "Dmitry Shkolnik @dshkol | Data: Census 2016, Statistics Canada")
  
  csd_div_plot <- ggplot(cma.csd) + 
    geom_sf(aes(fill = Ei, colour = Ei)) +  
    geom_sf(data = transit, size = 0.2, colour = "grey24") +
    geom_sf(data = roads, size = 0.2, colour = "grey36") +
    geom_sf(data = water, fill = "lightblue", colour = NA) + 
    coord_sf(datum = NA) +
    scale_fill_viridis_c("Diversity Entropy Index",
                         option = 3, breaks = c(0,0.5,1,1.5,2),
                         limits = c(0,2),
                         labels = c("Less\nDiverse","","","","More\nDiverse"),
                         guide = guide_legend(
                           direction = "horizontal",
                           title.position = "top",
                           label.position = "bottom",
                           keywidth = unit(2,"line"))) +
    scale_colour_viridis_c(option = 3, guide = "none",limits = c(0,2)) +
    theme(panel.background = element_blank(),
          legend.position = c(0.2,0.9),
          legend.background = element_blank(),
          legend.key = element_rect(color = NA)) +
    labs(caption = "Dmitry Shkolnik @dshkol | Data: Census 2016, Statistics Canada")
  objects <- list(ct = cma.ct, ct_plot = ct_div_plot,csd = cma.csd, csd_plot = csd_div_plot)
  return(objects)
}
```

#### Segregation

```{r segregation}

calc_h <- function(cma_obj) {
  cth <- cma_obj$ct %>% 
    select(GeoUID, CSD_UID, Population, Ei)
  st_geometry(cth) <- NULL
  
  cth <- cth %>%
    left_join(cma_obj$csd, by = c("CSD_UID"="GeoUID")) %>%
    select(GeoUID, CSD_UID, name, ctpop = Population.x,
           csdpop = Population.y, ctei = Ei.x, csdei = Ei.y) %>%
    group_by(GeoUID, CSD_UID) %>%
    filter(csdpop > 1000) %>%
    mutate(smallh = (ctpop*(csdei - ctei))/(csdei*csdpop)) %>%
    ungroup()

  csdh <- cth %>%
    group_by(CSD_UID, csdei) %>%
    summarise(H = sum(smallh, na.rm = TRUE)) %>% 
    right_join(cma_obj$csd,by = c("CSD_UID"="GeoUID"))
  
  return(csdh)
}
```


#### Import Data
```{r import}

library(dplyr)
library(sf)
library(cancensus) 
library(purrr)
library(xml2)
library(cancensusHelpers) # super specific helper functions 
library(ggplot2)
library(hrbrthemes)
library(rmapzen) # for vector tiles
library(ggrepel)

options(cancensus.api_key = "CensusMapper_5d949dee7f3e2546720b77d6ef08072e")

## Identifying the vectors for visible Minority status
parent_vector <- "v_CA16_3954"
minorities <- list_census_vectors("CA16") %>% 
  filter(vector == "v_CA16_3954") %>% 
  child_census_vectors(leaves_only = TRUE) %>% 
  pull(vector)

minority_vectors <- c(parent_vector, minorities)

# There are 244 ethnic origin vectors in the data. Let's get them at the CT level for the Vancouver metro region
```



#### Mapping diversity
```{r mapDiv}

diversity_csd_map <- function(csd) {
   
  csd.ct <- get_census("CA16", regions=list(CSD=csd), 
                       vectors = minority_vectors, level = "CT",
                       labels = "short", geo_format = NA)
  
  csd.csd <- get_census("CA16", regions=list(CSD=csd), 
                        vectors = minority_vectors, level = "CSD",
                        labels = "short", geo_format = NA)
  
  # Calculating diversity (Theil's E)
  # For every variable, divide by v_CA16_3999 and multiply times the logged inverse proportion, then
  # take the sum for each tract. With 14 different groups, the max entropy is ln(14) = 2.64
  base_pop <- quo(v_CA16_3954)
  csd.ei <- csd.ct %>% 
    group_by(GeoUID) %>% 
    mutate_at(minorities, funs(E = (./!!(base_pop))*(log(!!(base_pop)/.)))) %>%
    select(GeoUID, ends_with("_E")) %>% 
    ungroup() %>% 
    mutate_at(vars(ends_with("_E")), funs(ifelse(is.nan(.),0,.))) %>% 
    mutate(Ei = rowSums(select(.,-1), na.rm = FALSE)) %>% 
    select(GeoUID, Ei)
  
  csd.eicsd <- csd.csd %>% 
    group_by(GeoUID) %>% 
    mutate_at(minorities, funs(E = (./!!(base_pop))*(log(!!(base_pop)/.)))) %>%
    select(GeoUID, ends_with("_E")) %>% 
    ungroup() %>% 
    mutate_at(vars(ends_with("_E")), funs(ifelse(is.nan(.),0,.))) %>% 
    mutate(Ei = rowSums(select(.,-1), na.rm = FALSE)) %>% 
    select(GeoUID, Ei)
  
  # Join with geography
  
  csd.geo <- get_census_geometry("CA16", regions=list(CSD=csd), 
                                 level = "CT", geo_format = "sf")
  
  csd.csd.geo <- get_census_geometry("CA16", regions=list(CSD=csd), 
                                     level = "CSD", geo_format = "sf")
  
  # Which Census Tracts are in which CSD
  
  csd.ct <- csd.geo %>% 
    left_join(csd.ei) %>% 
    mutate()
  
  csd.csd <- csd.csd.geo %>% 
    left_join(csd.eicsd) %>% 
    mutate()
  
  # Adding map detail
  mz_set_tile_host_nextzen()
  get_vector_tiles <- function(bbox){
    mz_set_tile_host_nextzen(getOption("nextzen_API_key"))
    mx_box=mz_rect(bbox$xmin,bbox$ymin,bbox$xmax,bbox$ymax)
    mz_vector_tiles(mx_box, height = 1000, width = 1000)
  }
  
  bbox <- st_bbox(csd.ct)
  vector_tiles <- get_vector_tiles(bbox)
  
  if(length(vector_tiles$water$features) > 0) {
    water <- as_sf(vector_tiles$water)
  } 
  
  if(length(vector_tiles$roads$features) > 0) {
    roads <- as_sf(vector_tiles$roads) %>% 
      filter(kind == "highway")
  } else {roads <- water}
  
  if(length(vector_tiles$transit$features) > 0) {
    transit <- as_sf(vector_tiles$transit) %>% filter(kind == "subway")
  } else {transit <- water}
  
  ct_div_plot <- ggplot(csd.ct) + 
    geom_sf(aes(fill = Ei, colour = Ei)) +  
    geom_sf(data = csd.csd.geo, fill = NA, colour = "white") +
    geom_sf(data = transit, size = 0.2, colour = "grey24") +
    geom_sf(data = roads, size = 0.2, colour = "grey36") +
    geom_sf(data = water, fill = "lightblue", colour = NA) + 
    coord_sf(datum = NA) +
    scale_fill_viridis_c("Diversity Entropy Index",
                         option = 3, breaks = c(0,0.5,1,1.5,2), 
                         limits = c(0,2),
                         labels = c("Less\nDiverse","","","","More\nDiverse"),
                         guide = guide_legend(
                           direction = "horizontal",
                           title.position = "top",
                           label.position = "bottom",
                           keywidth = unit(2,"line"))) +
    scale_colour_viridis_c(option = 3, guide = "none",limits = c(0,2)) +
    theme(panel.background = element_blank(),
          legend.position = c(0.2,0.9),
          legend.background = element_blank(),
          legend.key = element_rect(color = NA)) + 
    labs(caption = "Dmitry Shkolnik @dshkol | Data: Census 2016, Statistics Canada")
  return(ct_div_plot)
}

# A function to calculate diversity scores only for CSD
diversity_csd <- function(cma) {
  cma.csd <- get_census("CA16", regions=list(CMA=cma), 
                        vectors = minority_vectors, level = "CSD",
                        labels = "short", geo_format = NA)
  
  # Calculating diversity (Theil's E)
  # For every variable, divide by v_CA16_3999 and multiply times the logged inverse proportion, then
  # take the sum for each tract. With 14 different groups, the max entropy is ln(14) = 2.64
  base_pop <- quo(v_CA16_3954)
  cma.eicsd <- cma.csd %>% 
    group_by(GeoUID,`Region Name`, Population) %>% 
    mutate_at(minorities, funs(E = (./!!(base_pop))*(log(!!(base_pop)/.)))) %>%
    select(GeoUID, `Region Name`, Population, ends_with("_E")) %>% 
    ungroup() %>% 
    mutate_at(vars(ends_with("_E")), funs(ifelse(is.nan(.),0,.))) %>% 
    mutate(Ei = rowSums(select(.,-c(1,2,3)), na.rm = FALSE)) %>% 
    mutate(CMA = cma) %>% 
    select(CMA, GeoUID, `Region Name`, Population, Ei)
  return(cma.eicsd)
}

# Apply function to get EI for all CSD for all Canadian CMAS
cmas <- list_census_regions("CA16") %>% filter(level == "CMA") %>% select(region, name, pop)

cma_ei <- purrr::map_df(cmas$region, .f = diversity_csd) %>% 
  left_join(cmas, by = c("CMA"="region")) %>% 
  select(`Region Name`, `CMA Name`=name, CMA, GeoUID, Population, `CMA Population` = pop, Ei)

# Top diverse CMAs with population over 5000

cma_ei %>% filter(Population > 5000) %>% arrange(desc(Ei)) %>% clean_names

# Pregenerating some sample CMA diversity objects - these make for the cool maps.

vancouver <- diversity_index("59933")
toronto <- diversity_index("35535")
montreal <-diversity_index("24462")
#ottawa <-diversity_index("505")
#quebeccity <- diversity_index("24421")
#winnipeg <-diversity_index("46602")
calgary <-diversity_index("48825")
#hamilton <-diversity_index("35537")
#stjohns <- diversity_index("10001")
#victoria <- diversity_index("59935")
#saguenay <- diversity_index("24408")
#abby <- diversity_index("59932")
#sask <- diversity_index("47725")
```


```{r mapSeg}
segregation_csd <- function(cma) {
  cma.ct <- get_census("CA16", regions=list(CMA=cma), 
                        vectors = minority_vectors, level = "CT",
                        labels = "short", geo_format = "sf")
  st_geometry(cma.ct) <- NULL
  
  cma.csd <- get_census("CA16", regions=list(CMA=cma), 
                        vectors = minority_vectors, level = "CSD",
                        labels = "short", geo_format = NA)
  
  base_pop <- quo(v_CA16_3954)
  
  cma.eict <- cma.ct %>% 
    group_by(GeoUID,`Region Name`, Population) %>% 
    mutate_at(minorities, funs(E = (./!!(base_pop))*(log(!!(base_pop)/.)))) %>%
    ungroup() %>% 
    select(GeoUID, CSD_UID, Population, ends_with("_E")) %>% 
    mutate_at(vars(ends_with("_E")), funs(ifelse(is.nan(.),0,.))) %>% 
    mutate(Ei = rowSums(select(.,-c(1,2,3)), na.rm = FALSE)) %>% 
    ungroup() %>% 
    select(GeoUID, CSD_UID, Population, Ei)

  cma.eicsd <- cma.csd %>%
    group_by(GeoUID,`Region Name`, Population) %>%
    mutate_at(minorities, funs(E = (./!!(base_pop))*(log(!!(base_pop)/.)))) %>%
    ungroup() %>%
    select(GeoUID, `Region Name`, Population, ends_with("_E")) %>%
    mutate_at(vars(ends_with("_E")), funs(ifelse(is.nan(.),0,.))) %>%
    mutate(Ei = rowSums(select(.,-c(1,2,3)), na.rm = FALSE)) %>%
    mutate(CMA = cma) %>%
    select(CMA, GeoUID, `Region Name`, Population, Ei)

  cma.h <- cma.eict %>%
    left_join(cma.eicsd, by = c("CSD_UID"="GeoUID")) %>%
    select(GeoUID, CSD_UID, name = `Region Name`, ctpop = Population.x,
           csdpop = Population.y, ctei = Ei.x, csdei = Ei.y) %>%
    filter(csdpop > 1000) %>%
    group_by(GeoUID, CSD_UID) %>%
    mutate(smallh = (ctpop*(csdei - ctei))/(csdei*csdpop)) %>%
    ungroup() %>%
    group_by(CSD_UID, csdei, name, csdpop) %>%
    summarise(H = sum(smallh, na.rm = TRUE)) %>% 
    mutate(cma = cma)
return(cma.h)
}

cma_seg <- purrr::map_df(cmas$region, .f = segregation_csd) %>% 
  left_join(cmas, by = c("cma"="region")) %>% 
  select(`Region Name`=name.x, `CMA Name`=name.y, CMA = cma,Population = csdpop, `CMA Population` = pop, Ei = csdei, H) %>% 
  ungroup()


clean_names <- function (dfr) {
    dfr <- dfr %>% mutate(name = as.character(name))
    replacement <- dfr %>% mutate(name = gsub(" \\(.*\\)", 
        "", name)) %>% pull(name)
    duplicated_rows <- c(which(duplicated(replacement, fromLast = TRUE)), 
        which(duplicated(replacement, fromLast = FALSE)))
    replacement[duplicated_rows] <- dfr$name[duplicated_rows]
    dfr$name <- factor(replacement)
    dfr
}

clean_names2 <- function (dfr) {
    dfr <- dfr %>% mutate(`Region Name` = as.character(`Region Name`))
    replacement <- dfr %>% mutate(`Region Name` = gsub(" \\(.*\\)", 
        "", `Region Name`)) %>% pull(`Region Name`)
    duplicated_rows <- c(which(duplicated(replacement, fromLast = TRUE)), 
        which(duplicated(replacement, fromLast = FALSE)))
    replacement[duplicated_rows] <- dfr$`Region Name`[duplicated_rows]
    dfr$`Region Name` <- factor(replacement)
    dfr
}
```

