---
title: "Retrieving Twitter data from its API"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(rtweet)
library(tidyverse)
library(knitr)
```

# First steps

Throughout this document we'll use the `rtweet` package [(link)](http://rtweet.info/index.html) to retrieve Twitter data from its API. Its use is quite straightforward once the authentication steps are completed...

## App creation 

All instructions on how to set up application and oauth are gathered here: (http://rtweet.info/articles/auth.html)

## authentication

I have stored my secret access info in a git-ignored script...

```{r}
source("twitter_token_creation.R")
twitter_token <- create_token(
    app =my_app ,
    consumer_key = my_API_key,
    consumer_secret = my_API_secret,
    access_token=my_access_token,
    access_secret=my_access_token_secret
)
```

Now we are ready to start gathering twitter info

# Search tweets

We can access tweets published during the last 7 days based on a keywords search (you can find out details about search rules [here](https://twitterdev.github.io/do_more_with_twitter_data/finding_the_right_data.html#finding-the-right-data): 

```{r}
tib_tweets <- search_tweets("(flat OR house) (sale OR sell OR selling)", n=500)
dim(tib_tweets)
```

There are no less than 88 variables related to these tweets.

The number of tweet returned by default is 100...  You can get many more than that if you settle on a broad keywords search... Beware of not exceeding the rate limit, though!

The object returned by `search_tweets()` is actually a *nested* tibble, that allows for multiple information to be stored for *one* tweet (i.e. multiple urls , multiple mentions of Twitter users in a single tweet, etc.): this means that some columns are not *vectors*, but *lists*:

```{r tweets_str}
str(tib_tweets,max.level=1)
```

I select some relevant, not-too-sparse variables just to display the kind of data we get out of `search_tweets()`

```{r}
tib_tweets %>%
  filter(lang=="en") %>% 
  select(screen_name,
         text,
         created_at,
         location)
```

```{r}
map_lgl(tib_tweets$geo_coords,~is.na(.x[1])) %>% 
  table()
```

The tweets are very seldomly located (variable `geo_coords`). On the other hand Twitter accounts often come with some information regarding location:

```{r}
tib_tweets %>% 
  select(location) %>% 
  sample_n(30)
```

These character strings related to location *might* or *might not* really make sense as geographical data... Anyway we can run some geocoding function to try and make them correspond to geographical coordinates:


Here I use Data Science Toolkit ("dsk") as source:

```{r}
library(ggmap)
geocode("NY",source="dsk", output="all")
```

The direct result of a call to function `geocode()` is quite messy so I tailored a function that extracts just the few pieces of information I need as a table:


```{r}
completeWithCoords=function(df){
  locs=data.frame(location=unique(df$location),
                  stringsAsFactors=FALSE)
  locs=locs$location
  locations=locs %>% 
    map(safely(geocode),source="dsk", output="all") %>% 
    map("result") %>% 
    map("results")
  coords=locations %>% 
    map(1) %>% 
    map("geometry") %>% 
    map("location") %>% 
    map(unlist) %>% 
    map(function(x){if(is.null(x)) x=c(lng=NA,lat=NA) else x=x})
  coords=do.call(rbind,coords)
  
  comp=locations %>% 
    map(1) %>% 
    map("address_components") %>% 
    map(function(x){map(x,safely(as.data.frame),stringsAsFactors=FALSE )}) %>% 
    map(function(x) map(x,"result")) %>% 
    map(function(x){do.call(rbind,x)})
  country=comp %>% 
    map(safely(function(x){filter(x,types=="country")})) %>% 
    map("result") %>% 
    map("long_name") %>% 
    map(function(x){if(is.null(x)) x=NA else x=x}) %>% 
    unlist()
  locality=comp %>% 
    map(safely(function(x){filter(x,types=="locality")}
    )) %>% 
    map("result") %>% 
    map("short_name") %>% 
    map(function(x){if(is.null(x)) x=NA else x=x}) %>% 
    unlist()
  area=comp %>% 
    map(safely(function(x){filter(x,types=="administrative_area_level_1")}
    )) %>% 
    map("result") %>% 
    map("short_name") %>% 
    map(function(x){if(is.null(x)) x=NA else x=x}) %>% 
    unlist()
  coords=data.frame(location=locs,
                    coords,
                    country,
                    locality,
                    area,
                    stringsAsFactors=FALSE)
  df=left_join(df,coords,by="location")
  return(df)
}
```

This function takes a table with column `location` and completes it with coordinates (latitude-longitude), country, locality, area.

```{r}
tib_trial <- tibble(location=c("NY",
                               "NYC",
                               "California, baby!",
                               "Midland, TX",
                               "Port Harcourt, Nigeria",
                               "All around the world!",
                               "Montmartre",
                               "La butte Montmartre",
                               "Quartier Latin",
                               "20 rue Mérieux Lyon"))
completeWithCoords(tib_trial)
```


# Other types of retrievable info

## Profile info a a particular user, friends and followers

```{r}
str(lookup_users("realtyWW"))
```

```{r}
str(get_followers("realtyWW"))
```
```{r}
str(get_friends("realtyWW"))
```

## Timelines

We can retrieve the tweets in the timeline of one Twitter-user. In that case the information retrieval is not limited to a certain time-window, but is **limited to a certain number of tweets** of the timeline (up to 3200).

Here we just retrieve 100 (the default number) tweets of user @realtyWW:

```{r}
timeline <- get_timeline("@realtyWW")
dim(timeline)
```

Again, the number of variables is quite **overwhelming**:

```{r}
str(timeline, max.level=1)
```

```{r}
timeline %>% 
  select(screen_name,
         text,
         created_at)
```

## Trending topics in a certain place

```{r}
get_trends("Lyon") %>% 
  kable()
```




